---
title: "Draw 2016/2017 Plot Boundaries"
author: Maggie Church
date: "2024-07-17"
updated: "2025-02-24"
output: html_document
---

We have a shapefile containing the plot boundaries for the areas that HAPET typically surveys annually. However, the 2016-2017 dataset surveyed additional plots. We need to manually draw the boundaries for 2016-2017 plots. 

1. This script creates an RShiny app that opens a map of the 2016-2017 delineated ponds, and logs the coordinates for rectangles drawn within the app. I use this to eyeball plot boundaries. 
2. I manually create polygon objects from those coordinates
3. I combine polygons into an sf object of 2016-2017 plot boundaries and create a plot id
4. Add these hand-drawn 2016-2017 plots to the usual plot boundaries
5. Export to data/allPlots

# Load libraries
```{r}
library(shiny)
library(leaflet)
library(leaflet.extras)
library(dplyr)
library(sf)
library(here)

wgs <- 4326
```

# Read in data
```{r}
# read in 2016-2017 HAPET data
pair16 <- read_sf(here("data", "raw_hapet", "Digitized_Wetlands", "2014_2017", "16PairsWAFinished.shp")) %>% st_transform(wgs)
brood16 <- read_sf(here("data", "raw_hapet", "Digitized_Wetlands", "2014_2017", "16BroodsWAFinished.shp")) %>% st_transform(wgs)
pair17 <- read_sf(here("data", "raw_hapet", "Digitized_Wetlands", "2014_2017", "17PairsWAFinished.shp")) %>% st_transform(wgs)
brood17 <- read_sf(here("data", "raw_hapet", "Digitized_Wetlands", "2014_2017", "17BroodsWAFinishedSurveyed.shp")) %>% st_transform(wgs)

# combine
needs_plots <- bind_rows(pair16, brood16, pair17, brood17)

# read in 2019-2024 plot boundaries
plots <- st_read(dsn = here("data", "raw_hapet", "FWS_2018_2019", "FSMS_Enterprise_Albers.gdb"), layer='FSMS_PLOTS') %>% st_transform(wgs)
```

# 1. Open a UI to streamline plot delineation
```{r, eval=F}
ui <- fluidPage(
  leafletOutput("map", height = "1000px"),  # Set the height to 600px
  verbatimTextOutput("bbox_data")  # Display extracted bounding box coordinates
)

server <- function(input, output, session) {
  
output$map <- renderLeaflet({
  leaflet() %>%
    addTiles() %>%
    addPolygons(data = needs_plots, fillColor = "blue", color = "black", weight = 1, opacity = 0.5) %>%
    addDrawToolbar(
      targetGroup = "drawn",
      rectangleOptions = TRUE,  
      polygonOptions = TRUE,   
      markerOptions = FALSE,
      circleOptions = FALSE,
      editOptions = editToolbarOptions()
    ) %>%
    addLayersControl(
      overlayGroups = c("drawn"),
      options = layersControlOptions(collapsed = FALSE)
    )
})

# Capture and extract bounding box coordinates when a rectangle is drawn
  observeEvent(input$map_draw_new_feature, {
      feature <- input$map_draw_new_feature
      
      # Extract the coordinates from the geometry of the drawn rectangle
      coords <- feature$geometry$coordinates[[1]]
      
      # Convert the list of coordinates to a matrix
      coords_matrix <- do.call(rbind, lapply(coords, unlist))
      
      # Extract the bounding box coordinates (min/max lat/lon)
      lng_min <- min(coords_matrix[, 1])  # Min longitude (x)
      lng_max <- max(coords_matrix[, 1])  # Max longitude (x)
      lat_min <- min(coords_matrix[, 2])  # Min latitude (y)
      lat_max <- max(coords_matrix[, 2])  # Max latitude (y)
      
      # Display the bounding box coordinates
      #bbox_coords <- sprintf("Bounding Box:\nLng min: %f, Lng max: %f\nLat min: %f, Lat max: %f", 
      #                       lng_min, lng_max, lat_min, lat_max)
      
     bbox_coords <- sprintf("%f, %f,  %f, %f,  %f, %f,  %f, %f,  %f, %f", 
                            lng_min, lat_min, lng_min, lat_max, 
                            lng_max, lat_max, lng_max, lat_min, 
                            lng_min, lat_min)
      
      output$bbox_data <- renderText({
        bbox_coords
      })
  })
}

shinyApp(ui, server)
```

# 2. Add plot bbox coordinates from the UI here
Note, I did miss some small ponds near plot boundaries, that's ok.
```{r}
# plot
p1 <- list(matrix(c(
-101.593838, 47.876980,  -101.593838, 47.906044,  -101.550922, 47.906044,  -101.550922, 47.876980,  -101.593838, 47.876980
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p2 <- list(matrix(c(
-101.572380, 47.964008,  -101.572380, 47.992794,  -101.529293, 47.992794,  -101.529293, 47.964008,  -101.572380, 47.964008
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p3 <- list(matrix(c(
-101.679153, 47.978317,  -101.679153, 48.007094,  -101.636066, 48.007094,  -101.636066, 47.978317,  -101.679153, 47.978317
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p4 <- list(matrix(c(
-101.641817, 48.052038,  -101.641817, 48.081061,  -101.598558, 48.081061,  -101.598558, 48.052038,  -101.641817, 48.052038
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p5 <- list(matrix(c(
-101.706791, 48.052267,  -101.706791, 48.080831,  -101.663189, 48.080831,  -101.663189, 48.052267,  -101.706791, 48.052267
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p6 <- list(matrix(c(
-101.749535, 48.052267,  -101.749535, 48.081290,  -101.706619, 48.081290,  -101.706619, 48.052267,  -101.749535, 48.052267
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p7 <- list(matrix(c(
-101.857080, 48.023285,  -101.857080, 48.052038,  -101.814466, 48.052038,  -101.814466, 48.023285,  -101.857080, 48.023285
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p8 <- list(matrix(c(
-101.728420, 48.182856,  -101.728420, 48.210947,  -101.685290, 48.210947,  -101.685290, 48.182856,  -101.728420, 48.182856
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p9 <- list(matrix(c(
-101.900768, 48.254627,  -101.900768, 48.283650,  -101.858196, 48.283650,  -101.858196, 48.254627,  -101.900768, 48.254627
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p10 <- list(matrix(c(
-102.105818, 48.356306,  -102.105818, 48.386867,  -102.062645, 48.386867,  -102.062645, 48.356306,  -102.105818, 48.356306
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p11 <- list(matrix(c(
-102.202721, 48.312827,  -102.202721, 48.341475,  -102.160406, 48.341475,  -102.160406, 48.312827,  -102.202721, 48.312827
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p12 <- list(matrix(c(
-102.245722, 48.341817,  -102.245722, 48.372444,  -102.202721, 48.372444,  -102.202721, 48.341817,  -102.245722, 48.341817
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p13 <- list(matrix(c(
-102.309408, 48.270169,  -102.309408, 48.296328,  -102.268124, 48.296328,  -102.268124, 48.270169,  -102.309408, 48.270169
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p14 <- list(matrix(c(
-102.289581, 48.341760,  -102.289581, 48.372615,  -102.246237, 48.372615,  -102.246237, 48.341760,  -102.289581, 48.341760
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p15 <- list(matrix(c(
-102.267694, 48.226102,  -102.267694, 48.254513,  -102.224607, 48.254513,  -102.224607, 48.226102,  -102.267694, 48.226102
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p16 <- list(matrix(c(
-102.138262, 48.124622,  -102.138262, 48.153375,  -102.095089, 48.153375,  -102.095089, 48.124622,  -102.138262, 48.124622
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p17 <- list(matrix(c(
-102.267609, 48.139631,  -102.267609, 48.167574,  -102.224607, 48.167574,  -102.224607, 48.139631,  -102.267609, 48.139631
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p18 <- list(matrix(c(
-102.353010, 48.110297,  -102.353010, 48.139001,  -102.310009, 48.139001,  -102.310009, 48.110297,  -102.353010, 48.110297
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p19 <- list(matrix(c(
-102.439699, 48.067182,  -102.439699, 48.095624,  -102.396955, 48.095624,  -102.396955, 48.067182,  -102.439699, 48.067182
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p20 <- list(matrix(c(
-102.547159, 48.226102,  -102.547159, 48.254741,  -102.504673, 48.254741,  -102.504673, 48.226102,  -102.547159, 48.226102
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p21 <- list(matrix(c(
-102.698650, 48.211977,  -102.698650, 48.240509,  -102.656593, 48.240509,  -102.656593, 48.211977,  -102.698650, 48.211977
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p22 <- list(matrix(c(
-102.698650, 48.153890,  -102.698650, 48.182455,  -102.656078, 48.182455,  -102.656078, 48.153890,  -102.698650, 48.153890
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p23 <- list(matrix(c(
-102.212505, 48.488510,  -102.212505, 48.517571,  -102.168989, 48.517571,  -102.168989, 48.488510,  -102.212505, 48.488510
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p24 <- list(matrix(c(
-102.255850, 48.444917,  -102.255850, 48.473832,  -102.212591, 48.473832,  -102.212591, 48.444917,  -102.255850, 48.444917
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p25 <- list(matrix(c(
-102.342367, 48.474116,  -102.342367, 48.503412,  -102.299538, 48.503412,  -102.299538, 48.474116,  -102.342367, 48.474116
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p26 <- list(matrix(c(
-102.386312, 48.459718,  -102.386312, 48.488681,  -102.342539, 48.488681,  -102.342539, 48.459718,  -102.386312, 48.459718
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p27 <- list(matrix(c(
-102.408028, 48.502673,  -102.408028, 48.532066,  -102.364769, 48.532066,  -102.364769, 48.502673,  -102.408028, 48.502673
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p28 <- list(matrix(c(
-102.364254, 48.517514,  -102.364254, 48.546217,  -102.321339, 48.546217,  -102.321339, 48.517514,  -102.364254, 48.517514
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p29 <- list(matrix(c(
-102.321253, 48.517457,  -102.321253, 48.546331,  -102.278080, 48.546331,  -102.278080, 48.517457,  -102.321253, 48.517457
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p30 <- list(matrix(c(
-102.342968, 48.560875,  -102.342968, 48.590121,  -102.299795, 48.590121,  -102.299795, 48.560875,  -102.342968, 48.560875
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p31 <- list(matrix(c(
-102.299881, 48.575358,  -102.299881, 48.602325,  -102.255850, 48.602325,  -102.255850, 48.575358,  -102.299881, 48.575358
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p32 <- list(matrix(c(
-102.386055, 48.589837,  -102.386055, 48.618612,  -102.343140, 48.618612,  -102.343140, 48.589837,  -102.386055, 48.589837
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p33 <- list(matrix(c(
-102.429657, 48.590007,  -102.429657, 48.618612,  -102.386312, 48.618612,  -102.386312, 48.590007,  -102.429657, 48.590007
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p34 <- list(matrix(c(
-102.429743, 48.560988,  -102.429743, 48.589780,  -102.386312, 48.589780,  -102.386312, 48.560988,  -102.429743, 48.560988
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p35 <- list(matrix(c(
-102.407513, 48.633533,  -102.407513, 48.662170,  -102.364426, 48.662170,  -102.364426, 48.633533,  -102.407513, 48.633533
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p36 <- list(matrix(c(
-102.515659, 48.618953,  -102.515659, 48.647995,  -102.473516, 48.647995,  -102.473516, 48.618953,  -102.515659, 48.618953
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p37 <- list(matrix(c(
-102.581663, 48.619066,  -102.581663, 48.647938,  -102.538061, 48.647938,  -102.538061, 48.619066,  -102.581663, 48.619066
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p38 <- list(matrix(c(
-102.537718, 48.662283,  -102.537718, 48.691357,  -102.494631, 48.691357,  -102.494631, 48.662283,  -102.537718, 48.662283
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p39 <- list(matrix(c(
-102.646894, 48.619066,  -102.646894, 48.647825,  -102.603979, 48.647825,  -102.603979, 48.619066,  -102.646894, 48.619066
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p40 <- list(matrix(c(
-102.712469, 48.619009,  -102.712469, 48.647655,  -102.668953, 48.647655,  -102.668953, 48.619009,  -102.712469, 48.619009
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p41 <- list(matrix(c(
-102.712297, 48.662396,  -102.712297, 48.691187,  -102.668695, 48.691187,  -102.668695, 48.662396,  -102.712297, 48.662396
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p42 <- list(matrix(c(
-102.721395, 48.735021,  -102.721395, 48.764223,  -102.677450, 48.764223,  -102.677450, 48.735021,  -102.721395, 48.735021
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p43 <- list(matrix(c(
-102.742510, 48.764110,  -102.742510, 48.792277,  -102.700024, 48.792277,  -102.700024, 48.764110,  -102.742510, 48.764110
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p44 <- list(matrix(c(
-102.787142, 48.749624,  -102.787142, 48.778365,  -102.742767, 48.778365,  -102.742767, 48.749624,  -102.787142, 48.749624
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p45 <- list(matrix(c(
-102.851086, 48.720244,  -102.851086, 48.749228,  -102.809029, 48.749228,  -102.809029, 48.720244,  -102.851086, 48.720244
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p46 <- list(matrix(c(
-102.895546, 48.705859,  -102.895546, 48.734795,  -102.852716, 48.734795,  -102.852716, 48.705859,  -102.895546, 48.705859
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p47 <- list(matrix(c(
-102.939663, 48.749455,  -102.939663, 48.778478,  -102.896147, 48.778478,  -102.896147, 48.749455,  -102.939663, 48.749455
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p48 <- list(matrix(c(
-102.821217, 48.633306,  -102.821217, 48.662283,  -102.777357, 48.662283,  -102.777357, 48.633306,  -102.821217, 48.633306
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p49 <- list(matrix(c(
-102.754354, 48.590007,  -102.754354, 48.618612,  -102.713156, 48.618612,  -102.713156, 48.590007,  -102.754354, 48.590007
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p50 <- list(matrix(c(
-102.951164, 48.633759,  -102.951164, 48.662170,  -102.907562, 48.662170,  -102.907562, 48.633759,  -102.951164, 48.633759
    ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p51 <- list(matrix(c(
-103.080940, 48.589950,  -103.080940, 48.619009,  -103.037767, 48.619009,  -103.037767, 48.589950,  -103.080940, 48.589950
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p52 <- list(matrix(c(
-102.993221, 48.474458,  -102.993221, 48.498465,  -102.953653, 48.498465,  -102.953653, 48.474458,  -102.993221, 48.474458
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p53 <- list(matrix(c(
-102.515831, 48.503128,  -102.515831, 48.531669,  -102.473688, 48.531669,  -102.473688, 48.503128,  -102.515831, 48.503128
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p54 <- list(matrix(c(
-102.537117, 48.401172,  -102.537117, 48.430738,  -102.496090, 48.430738,  -102.496090, 48.401172,  -102.537117, 48.401172
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p55 <- list(matrix(c(
-102.647238, 48.356363,  -102.647238, 48.386354,  -102.601833, 48.386354,  -102.601833, 48.356363,  -102.647238, 48.356363
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p56 <- list(matrix(c(
-102.720966, 48.314825,  -102.720966, 48.341931,  -102.677622, 48.341931,  -102.677622, 48.314825,  -102.720966, 48.314825
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p57 <- list(matrix(c(
-102.711697, 48.387152,  -102.711697, 48.416100,  -102.669468, 48.416100,  -102.669468, 48.387152,  -102.711697, 48.387152
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p58 <- list(matrix(c(
-103.059654, 48.372900,  -103.059654, 48.401172,  -103.020859, 48.401172,  -103.020859, 48.372900,  -103.059654, 48.372900
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326)

# plot
p59 <- list(matrix(c(
-103.167114, 48.416556,  -103.167114, 48.442298,  -103.125229, 48.442298,  -103.125229, 48.416556,  -103.167114, 48.416556
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p60 <- list(matrix(c(
-103.166943, 48.676907,  -103.166943, 48.705576,  -103.124714, 48.705576,  -103.124714, 48.676907,  -103.166943, 48.676907
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p61 <- list(matrix(c(
-103.211060, 48.662623,  -103.211060, 48.691074,  -103.167973, 48.691074,  -103.167973, 48.662623,  -103.211060, 48.662623
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p62 <- list(matrix(c(
-103.254662, 48.662510,  -103.254662, 48.691074,  -103.212261, 48.691074,  -103.212261, 48.662510,  -103.254662, 48.662510
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p63 <- list(matrix(c(
-103.245907, 48.706256,  -103.245907, 48.734795,  -103.201447, 48.734795,  -103.201447, 48.706256,  -103.245907, 48.706256
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p64 <- list(matrix(c(
-103.180161, 48.735814,  -103.180161, 48.763997,  -103.136902, 48.763997,  -103.136902, 48.735814,  -103.180161, 48.735814
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p65 <- list(matrix(c(
-103.246250, 48.749738,  -103.246250, 48.778478,  -103.202133, 48.778478,  -103.202133, 48.749738,  -103.246250, 48.749738
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p66 <- list(matrix(c(
-103.136559, 48.793182,  -103.136559, 48.821446,  -103.092957, 48.821446,  -103.092957, 48.793182,  -103.136559, 48.793182
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p67 <- list(matrix(c(
-103.319550, 48.663190,  -103.319550, 48.691074,  -103.276463, 48.691074,  -103.276463, 48.663190,  -103.319550, 48.663190
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p68 <- list(matrix(c(
-103.552494, 48.851388,  -103.552494, 48.879845,  -103.509407, 48.879845,  -103.509407, 48.851388,  -103.552494, 48.851388
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p69 <- list(matrix(c(
-103.661327, 48.880522,  -103.661327, 48.908624,  -103.619614, 48.908624,  -103.619614, 48.880522,  -103.661327, 48.880522
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p70 <- list(matrix(c(
-103.760462, 48.938344,  -103.760462, 48.962920,  -103.728104, 48.962920,  -103.728104, 48.938344,  -103.760462, 48.938344
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p71 <- list(matrix(c(
-103.793764, 48.967372,  -103.793764, 48.996044,  -103.750162, 48.996044,  -103.750162, 48.967372,  -103.793764, 48.967372
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p72 <- list(matrix(c(
-103.837452, 48.923909,  -103.837452, 48.952663,  -103.795996, 48.952663,  -103.795996, 48.923909,  -103.837452, 48.923909
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p73 <- list(matrix(c(
-103.946800, 48.938570,  -103.946800, 48.966864,  -103.903799, 48.966864,  -103.903799, 48.938570,  -103.946800, 48.938570
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p74 <- list(matrix(c(
-103.705444, 48.733323,  -103.705444, 48.763884,  -103.662872, 48.763884,  -103.662872, 48.733323,  -103.705444, 48.733323
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p75 <- list(matrix(c(
-103.814621, 48.792616,  -103.814621, 48.821333,  -103.772049, 48.821333,  -103.772049, 48.792616,  -103.814621, 48.792616
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p76 <- list(matrix(c(
-103.880711, 48.822350,  -103.880711, 48.851444,  -103.837280, 48.851444,  -103.837280, 48.822350,  -103.880711, 48.822350
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 

# plot
p77 <- list(matrix(c(
-104.266090, 48.896719,  -104.266090, 48.924078,  -104.220943, 48.924078,  -104.220943, 48.896719,  -104.266090, 48.896719
      ), ncol = 2, byrow = TRUE)) %>% 
  st_polygon() %>%
  st_sfc(crs = 4326) 
```

# 3. Combine hand-drawn 2016-2017 plots and create plotid
```{r}
# Store all polygons in a list
polygons_list <- list(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12, p13, p14, p15, p16, p17, p18, p19, p20, p21, p22, p23, p24, p25, p26, p27, p28, p29, p30, p31, p32, p33, p34, p35, p36, p37, p38, p39, p40, p41, p42, p43, p44, p45, p46, p47, p48, p49, p50, p51, p52, p53, p54, p55, p56, p57, p58, p59, p60, p61, p62, p63, p64, p65, p66, p67, p68, p69, p70, p71, p72, p73, p74, p75, p76, p77)

# get max plot id from HAPET
max_hapet_plotid <- plots %>% st_drop_geometry() %>% summarize(max(Plot, na.rm=T)) %>% pull()

# create new plot ids, atop the max id from HAPET 
plot_ids <- 1:length(polygons_list) + max_hapet_plotid

# Extract the actual polygon geometries from each sfc object
plots_1617 <- lapply(polygons_list, function(p) p[[1]]) %>%
  st_sfc(crs = 4326) %>% 
  st_sf(Plot = plot_ids,
        geometry = .) 
```

# 4. Add these 2016-2017 boundaries to the 2019-2024 boundaries shp
```{r}
plots_all <- plots %>% rename(geometry=Shape) %>% bind_rows(plots_1617)
```

# 5. Write plot boundaries to file
```{r}
st_write(plots_all, here("data", "allPlots", "allPlots.shp"), append=FALSE)
```

