---
title: "Combine Raw HAPET"
author: Maggie Church
created: "2024-07-17"
updated: "2026-02-17"
---

**Overview:** This script preps HAPET pond polygons for analysis

**Inputs:**

-   Plot boundary shapefile (created in script #1)

-   Raw HAPET pond shapefiles

**Steps:**

1.  Performs specific data-cleaning steps to different survey periods (2016-2017, 2019, 2020, 2021+)

2.  Combines the clean pond data into one shapefile

3.  Creates a training pond dataset: excludes polygons \< 400m2

4.  Creates a testing pond dataset: merges split polygons

**Outputs:**

-   allPond.shp (training ponds)

-   allPond_all.shp (testing ponds)

# Setup environment and load libraries
```{r}
library(tidyverse)
library(sf)
library(here)

set.seed(10)

# Uses UTM Zone 14N for spatial operations: st_intersects, st_area, st_join, st_touches (distortion outside of Zone 14N is negligible)
wgs <- 4326
utm14 <- 32614 

# Using this bc 2020 kept throwing issues. It loosens rules about geometric validity
sf_use_s2(FALSE)

Sys.setenv("SHAPE_RESTORE_SHX" = "YES")
```

# Read in plot boundaries (from script #1)

```{r}
# read in plot boundary data - I'll use this to drop entire plots from the 2019 dataset, and add a plot id column to the final combined dataset
plots <- st_read(here("data", "allPlots", "allPlots.shp")) |> st_transform(utm14)
```

# 1. Clean raw pond datasets

### 2016-2017 wet area
Read in the 2016 and 2017 data, stored in 4 separate shapefiles. Add flyover date and veg attributes. The following is from HAPET documentation: digitized via "an eCognition process and confirmed by the GIS intern" Plot_ID - identifies a 10.36 square km area in which wetlands were surveyed. Wet_Acres - identifies the wet acreage of each wetland (as digitized - yes, sorry not in metric). PW - ground-truthed observer estimate of percent of the basin that was wet. 0 = not ground-truthed Veg_Height - only present in the Pair (Spring) wet area files. Was not collected during brood surveys. PE - identifies the percentage of the basin that observers on the ground estimated was covered by emergent vegetation.
```{r, warning=F, message=F}
pair16 <- read_sf(here("data", "raw_hapet", "Digitized_Wetlands", "2014_2017", "16PairsWAFinished.shp")) |> st_transform(utm14)
brood16 <- read_sf(here("data", "raw_hapet", "Digitized_Wetlands", "2014_2017", "16BroodsWAFinished.shp")) |> st_transform(utm14)
pair17 <- read_sf(here("data", "raw_hapet", "Digitized_Wetlands", "2014_2017", "17PairsWAFinished.shp")) |> st_transform(utm14)
brood17 <- read_sf(here("data", "raw_hapet", "Digitized_Wetlands", "2014_2017", "17BroodsWAFinishedSurveyed.shp")) |> st_transform(utm14)

# read dates of aerial photography (note, not the on-the-ground data collection date)
aerialphotos <- read_sf(here("data", "raw_hapet", "Aerial_photos", "AllPlots.shp")) |>as.data.frame()

# add aerial survey date to the 2016 and 2017 data
# also, add the contrived plot number (created in 1.draw_1617plots.Rmd)
pair16 <- aerialphotos |>
  select(PLOT_ID, Date16_A) |>
  right_join(pair16, by = c('PLOT_ID'='Plot_ID')) |>
  st_as_sf() 

brood16 <- aerialphotos |>
  select(PLOT_ID, Date16_B) |>
  right_join(brood16, by = c('PLOT_ID'='Plot_ID')) |>
  st_as_sf() 

pair17 <- aerialphotos |>
  select(PLOT_ID, Date17_A) |>
  right_join(pair17, by = c('PLOT_ID'='Plot_ID')) |>
  st_as_sf() 

brood17 <- aerialphotos |>
  select(PLOT_ID, Date17_B) |>
  right_join(brood17, by = c('PLOT_ID'='Plot_ID')) |>
  st_as_sf() 
```

### 2019 wet area
Data note: The 2019 shapefile has some duplicated ponds where digitizations overflow into adjacent plots. I keep the largest of overlapping ponds. Anticipated public HAPET release should include QA/QC corrections.
```{r, warning=F, message=F}
# read in raw data
all19_raw <- read_sf(here("data", "raw_hapet", "FWS_2018_2019", "2019_shapefiles", "2019_ALL.shp")) |>
  st_transform(utm14) |>
  st_make_valid() |> 
  filter(st_coordinates(st_centroid(geometry))[, "Y"] > 30) # some weird ones near the equator...

# remove entire plots - I hand QAed these
all19_raw <- all19_raw |>
  st_join(select(plots, Plot)) |>
  filter(!(Plot %in% c(346, 609, 807, 223, 218)))  |>  # these weren't actually surveyed, they're just adjacent to plots that were
  filter(!(Plot %in% c(203)))                          # digitization of this one was too wonky

# Create a helper function to get intersecting pond IDs
get_intersecting_ids <- function(index, intersect_matrix, id_column) {
    intersecting_ids <- id_column[intersect_matrix[[index]]]
    intersecting_ids <- intersecting_ids[intersecting_ids != id_column[index]] # Exclude self
    paste(intersecting_ids, collapse = ",")
}

# Use st_intersects to find intersections
intersection_matrix <- st_intersects(all19_raw, sparse = TRUE)

# Extract IDs of intersecting polygons
all19_wint <- all19_raw |>
    mutate(intersecting_ids = 
             sapply(
                 seq_along(intersection_matrix), 
                 get_intersecting_ids, 
                 intersect_matrix = intersection_matrix, 
                 id_column = all19_raw$OBJECTID))

# max number of overlaps with a polygon
max_ints <- all19_wint |> 
  as.data.frame() |>
  mutate(num_intersections = str_count(intersecting_ids, ",")) |>
  summarize(max(num_intersections)) |>
  pull()

# some ponds have multiple overlaps. Break these into separate columns
all19_wint_sep <- all19_wint |>
  separate(intersecting_ids, 
           into = paste0('int', 1:(max_ints+1)), 
           sep = ",") 

# map of pond id <-> size
pond_sizes <- all19_raw |>
  as.data.frame() |>
  mutate(OBJECTID = as.character(OBJECTID)) |>
  select(OBJECTID, 
         intersecting_Shape_Area = Shape_Area) 

# add size of intersecting pond
all19_wint_long <- all19_wint_sep |>
  filter(int1 != '') |>
  mutate(id_og = OBJECTID) |>
  mutate(int0 = as.character(OBJECTID)) |>
  select(id_og, int0, int1:int17) |>
  pivot_longer(cols = c(int0, int1:int17), 
               names_to = "int", 
               values_to = "OBJECTID") |>
  filter(!is.na(OBJECTID)) |>
  left_join(pond_sizes) 

# identify the ponds that aren't the largest of overlaps
temp_not_largest <- all19_wint_long |>
  group_by(id_og) |>
  filter(intersecting_Shape_Area == max(intersecting_Shape_Area)) |>
  filter(id_og != OBJECTID) # these are the ones to be deleted bc theyre smaller

# remove ponds that aren't the largest of overlaps
all19 <- all19_raw |>
  filter(!OBJECTID %in% temp_not_largest$id_og)
```

### 2020 wet area
The 2020 shapefiles are stored separately, and have some geometry issues. (We're not going to use this probably - 2020 data came from Planet instead of aerial surveys bc of COVID)
```{r, warning=F, message=F}
# List all gdb files in the raw 2020 folder
files <-  list.files(here("data", "raw_hapet", "FWS_2020", "AllWMD2020Water"), full.names = TRUE)

# fn to convert multisurface features to multipolygon
ms_to_mp <- function(geom) {
  if (st_is(geom, "MULTISURFACE")) {
    geom <- st_cast(geom, "MULTIPOLYGON")
  }
  return(geom)
}

# initialize empty list, for storing output of loop
feat_list <- list()

# loop through raw data to store imported features
for(f in files) {
  
  # list layers in the gdb "f"
  layers <- st_layers(dsn = f)
  
  for(l in layers$name) {
    
    # print loop iteration
    print(paste(basename(f), l, sep = "_"))
    
    # read the layer
    feat <- st_read(dsn = f, layer=l) |>
      st_transform(utm14)
    
    # geometry is called Shape or SHAPE. Make uniform
    if ("SHAPE" %in% colnames(feat)) {feat <- rename(feat, Shape = SHAPE)}
    
    # Convert multisurface to multipolygon, if necessary
    feat <- feat |>
      mutate(geometry = st_sfc(lapply(st_geometry(feat), ms_to_mp))) |>
      st_set_geometry("geometry")  |>     # Set the new column as the geometry
      select(-Shape) |>                   # drop original geometry
      st_set_crs(utm14)                   # reset crs if it got dropped during st_set_geometry
    
    # Ensure unique column names
    colnames(feat) <- make.unique(colnames(feat))
    
    # store feature
    feat_list[[paste(basename(f), l, sep = "_")]] <- feat
  }
}

# combine features
all20 <- bind_rows(feat_list)

# Fix invalid geometries
all20 <- st_make_valid(all20)
all20 <- all20[st_geometry_type(all20) != "GEOMETRYCOLLECTION", ]

rm(feat)
rm(feat_list)
rm(layers)
```

### 2021-2024 wet area
This data is solid :)
```{r, warning=F, message=F}
read_21_24 <- function(filename){
  
  # get gdb layers
  path <- paste0(here("data", "raw_hapet", "2021_2024", filename))
  layers <- st_layers(path)
  
  # Read all layers into a list of sf objects
  sf_list <- 
    lapply(layers$name, function(layer) {
      st_read(path, layer = layer) |>
      st_transform(utm14) |>
      rename(geometry=Shape)
  })
  
  sf <- bind_rows(sf_list)
}

all21 <- read_21_24("WetnessNoOwner2021.gdb") |> st_cast("MULTIPOLYGON") 
all22 <- read_21_24("WetnessNoOwner2022.gdb") |> st_cast("MULTIPOLYGON") 
all23 <- read_21_24("WetnessNoOwner2023.gdb") |> st_cast("MULTIPOLYGON") 
all24 <- read_21_24("WetnessNoOwner2024.gdb") |> st_cast("MULTIPOLYGON") 
```

# 2. Combine years
```{r}
# combine datasets
allPonds <- 
  pair16 |>
  select(flyover_date = Date16_A) |>
  mutate(dataset = 'pair16') |>
  bind_rows(brood16 |>
              select(flyover_date = Date16_B, PE) |>
              mutate(dataset = 'brood16')) |>
  bind_rows(pair17 |>
              select(flyover_date = Date17_A) |>
              mutate(dataset = 'pair17')) |>
  bind_rows(brood17 |>
              select(flyover_date = Date17_B, PE) |>
              mutate(dataset = 'brood17')) |>
  bind_rows(all19 |>
              select() |>
              mutate(dataset = '2019')) |>
  bind_rows(all20 |>
              select() |>
              mutate(dataset = '2020')) |>
  bind_rows(all21 |> 
              select(flyover_date = Date, PercFull) |>
              mutate(dataset = '2021')) |>
  bind_rows(all22 |> 
              select(flyover_date = Date, PercFull) |>
              mutate(dataset = '2022')) |>
  bind_rows(all23 |> 
              select(flyover_date = Date, PercFull) |>
              mutate(dataset = '2023')) |>
  bind_rows(all24 |> 
              select(flyover_date = Date, PercFull) |>
              mutate(dataset = '2024')) |>
  filter(st_is_valid(geometry)) |>                    # Filter to keep only valid geometries
  mutate(area_m = as.numeric(st_area(geometry))) |>   # I know most have an area already, but I redo to ensure consistency
  st_join(select(plots, Plot))  |>                    # Add plot_id
  mutate(rownum = row_number())                       # Add rownum
```

Additional QA
```{r}
# ~300 ponds span multiple plots - split these ponds into 2+ by plot boundary
dup <- allPonds |>
  group_by(rownum) |>
  filter(n()>1) |>
  st_intersection(select(plots, Plot)) |>
  mutate(Plot=Plot.1) |>
  select(-Plot.1) |>
  distinct()

# add these de-duplicated ponds back into the full dataset (nuke rownum and redo it)
allPonds2 <- allPonds |>
  group_by(rownum) |>
  filter(n()==1) |>  # drop duplicated ponds
  ungroup() |>
  bind_rows(dup) |>
  select(-rownum) |>
  mutate(rownum=row_number())

# There are a handful of ponds that are somehow now point format. Just drop these.
allPonds3 <- allPonds2[st_geometry_type(allPonds2) %in% c("POLYGON", "MULTIPOLYGON"), ]

# oh and drop 2019 "overflowing" data
allPonds4 <- allPonds3 |> filter(!(is.na(Plot) & dataset=="2019"))
```

# 3. Export allPond dataset (used for training)
Filter to ponds \>= 400m2, for training dataset
I might have picked a more descriptive name if I had forseen that I'd need to save a separate shapefile for testing ponds (next chunk)
```{r}
# filter ponds < 20x20m
allPonds_gt400m2 <- allPonds4 |> filter(area_m >= 400)

st_write(allPonds_gt400m2, here("data", "allPonds", "allPond.shp"),  append=F)
```

# 4. Export allPond_all dataset (used for testing)
Here, I export all ponds (including ponds \< 400m2). I really only use this for testing.

Also: so, HAPET splits merged ponds (bc they have to use NWI polygons as reference). Below, I merge these split ponds back together. It's important to do for the test ponds because it affects pond counts and pond sizes.

I'm not going to merge ponds for allPonds_gt400m2, because I only realized this after I built the model. I don't want to mess with that dataset (for reproducibility), and I don't want to recreate the entire model. It shouldn't affect model performance anyways.

Note that HAPET made clean splits for 2019+, but added 1 pixel of separation for splits prior to 2019. Those are harder to address (so I don't).
```{r}
datasets <- allPonds4 %>% 
  st_drop_geometry() %>% # speeds up the next part
  distinct(dataset) %>% 
  pull()

# initialize empty vector
out <- vector("list", length(datasets))

for (i in seq_along(datasets)) {
  
  ponds_yr <- allPonds4 %>% 
    filter(dataset == datasets[i]) %>% 
    st_cast("POLYGON") 
  
  # Compute touches ONCE on the full dataset
  touch_list <- st_touches(ponds_yr)
  touch_counts <- lengths(touch_list)
  
  # Split dataset
  touching     <- ponds_yr[touch_counts > 0, ]
  non_touching <- ponds_yr[touch_counts == 0, ]
  
  # Build graph ONLY on touching polygons
  touch_graph <- st_touches(touching)
  g <- igraph::graph_from_adj_list(touch_graph)
  components <- igraph::components(g)$membership
  
  # Union touching polygons by connected component
  touching_union <- touching %>%
    mutate(group = components) %>%
    summarise(
      Plot = first(Plot),
      dataset = first(dataset),
      geometry = st_union(geometry),
      .groups = "drop"
    )
  
  # write
  out[[i]] <- bind_rows(touching_union, non_touching)
}

allPonds_clean <- bind_rows(out) %>% 
  mutate(rownum = row_number()) # re-do once more

st_write(allPonds_clean, here("data", "allPonds_all", "allPond_all.shp"),  append=F)
```