---
title: "Buffer Raw HAPET"
author: Maggie Church
created: "2025-02-24"
updated: "2025-02-24"
---

**Overview:**

This script prepares the pond polygons for training point selection: to avoid training in mixed pixels, we buffer ponds by 7.5m. The reason I created an entirely separate script for this step is because it takes a long time!

Inputs:

-   HAPET ponds shapefile (created in script #2 - this is the "training" version that drops ponds <400m2 and doesn't merge split ponds)

Steps

1.  In-buffer ponds by 7.5m, for wet point sampling

2.  Out-buffer ponds by 7.5, for dry point sampling

Outputs:

-   allPonds_inbuff75.shp

-   allPonds_outbuff75.shp

# Load libraries
```{r, message=F, warning=F, results='hide'}
library(tidyverse)
library(sf)
library(here)
```

# Read data
Read in preprocessed and combined HAPET data (this is the output of `2.combine_raw_hapet.Rmd`).
```{r}
# Load HAPET ponds
allPonds <- st_read(here("data", "allPonds", "allPond.shp"))
```

# 1. In-buffer ponds
Apply a 7.5m inward buffer to ponds, to ensure 10-m pixels are mostly wet (this takes a few minutes)
```{r}
# Remove pond edges
allPonds_inbuff75 <- 
  st_buffer(allPonds, -7.5) |> 
  filter(st_is_empty(geometry)==FALSE)  # drop ponds that get "buffered out"

# Export
st_write(allPonds_inbuff75, here("data", "intermediate", "buffered_ponds", "allPonds_inbuff75.shp"))
```

# 2. Out-buffer ponds
Apply a 7.5m outward buffer to ponds, to ensure 10-m pixels are mostly dry (this takes a few minutes)
```{r}
# Remove pond edges
allPonds_outbuff75 <- 
  st_buffer(allPonds, 7.5) |> 
  filter(st_is_empty(geometry)==FALSE)  # drop ponds that get "buffered out"

st_write(allPonds_outbuff75, here("data", "intermediate", "buffered_ponds", "allPonds_outbuff75.shp"))
```