---
title: "Process Possibility Rasters"
author: Maggie Church
created: "2024-09-24"
updated: "2025-02-24"
---

**Overview:** The "possibility" raster is what we call our historic wetland footprint raster, created in GEE using NWI, JRC's GSW, SSURGO, HAPET, and Census Roads. We use possible wetlands as our sampling ROI for dry points in script #7. However, the GEE-exported raster tiles are large and difficult to work with, so this script breaks them into individual plot-size rasters. This script is similar to script #4, except it also merges the tiles to export a single ppjv-wide possibility raster, which we use later for model evaluation.

**Inputs**:

-   Plot boundary shapefile (script #1)

-   PPJV administrative boundary

-   Possibility Layer raster tiles (link to GEE code that creates the Possibility layer [here](https://code.earthengine.google.com/05a27818e5e2a0e36c48739d8cebb1f9?noload=1){.uri})

**Steps**:

1. Merge raster tiles into a single ppjv-wide raster

2. Crop into 651 rasters - one for each plot

**Outputs:**

-   A PPJV-wide possibility raster, saved in `data/gee_exports/gee_rasters/possibilityLayer/`

-   651 possibility rasters, saved in `data/intermediate/rasterprocessing/cropped_poss_rasters`

**NOTE:** When I originally ran this, I'd exported the Possibility layer from GEE in wgs84 (epsg 4326), then reprojected into UTM14 and ran this code. The reprojection made the result continuous 0-1 instead of binary. That's what exists in data/intermediate/raster_processing/cropped_poss_rasters right now, bc I haven't re-run with this updated code. 

# Setup environment and load libraries
```{r}
library(sf)
library(dplyr)
library(terra)
library(future.apply)
library(parallel)
library(tools)
library(here)

utm14 <- 32614
utm14_str <- "+proj=utm +zone=14 +datum=WGS84 +units=m +no_defs"
w <- detectCores()

ppr <- st_read(here("data", "boundaries", "PPJV")) |> st_transform(utm14)
```

### 1. Process possibilityLayer tiles into 1 single raster
We use this layer a lot, so let's just merge and save it.
Run once! 
```{r}
# read in the list of possibility layer files (these are in UTM14)
tif_files <-  list.files(path=here("data", "gee_exports", "gee_rasters", "possibilityLayer"),
                      pattern = "\\.tif$",
                     full.names = TRUE)

tif_list <- lapply(tif_files, terra::rast)

# Combine into a single mosaic
mosaic_raster <- do.call(terra::mosaic, c(tif_list))

# mask and trim
mosaic_raster_trimmed <- mosaic_raster |>
  mask(ppr) |>
  trim()

output_path <- here(paste0("data/gee_exports/gee_rasters/possibilityLayer/"))

# save
writeRaster(mosaic_raster_trimmed,
            filename = paste0(output_path,"possibilityLayer_updated.tif"),
            overwrite = TRUE,
            gdal=c("COMPRESS=DEFLATE", "TFW=YES"),
            datatype='INT1U') # accepts integers 1-255 (smaller files than the default FLT4S)
```

# 2. Break (processed) possibility raster tiles into plot-size rasters

## Load plot boundaries 
Apply inward plot buffer, because we're not going to sample any training points at plot edges. This is quick.
```{r, message=F, warning=F}
# read in plot boundary data
plots <- 
  st_read(here("data", "allPlots")) |>
  st_transform(utm14)

# apply inwards plot buffer 
plots_inbuff75 <- st_buffer(plots, -7.5)
```

## Function to split tiles
This next chunk creates the function we'll use to crop the larger Possibility Layer into smaller plot-size tiles.
Unlike script #4, we merged the Possibility Layer tiles in step 1, so we don't need to account for split-tile plots now.
```{r}
crop_raster <- function(p, raster) {
  
  # Plot extent for the given plot ID
  plot_extent <- plots_inbuff75[plots_inbuff75$Plot == p, ]
  
  # Try cropping
  cropped <- tryCatch({
      crop(raster, plot_extent)
    },
    error = function(e) NULL
  )
  
  # If crop failed or contains no data, return NULL
  if (is.null(cropped) || all(is.na(values(cropped)))) {
    return(NULL)
  }
  
  # Define output file path
  output_file <- file.path(
    here("data", "intermediate", "raster_processing", "cropped_poss_rasters"),
    paste0("raster_", p, ".tif")
  )
  
  # Save cropped raster
  writeRaster(cropped, output_file, overwrite = TRUE)
  
  return(output_file)
}
```

## Run the function
Now we run the function! I use parallel processing to speed things up... this part only takes a few minutes.
Note the permanent lake layer is EPSG=32614 (UTM 14N)
```{r}
# Unique plot IDs
plot_ids <- unique(plots_inbuff75$Plot)

# Parallel plan
plan(multisession, workers = w - 2)

# Apply crop for each plot
result_files <- lapply(plot_ids, function(p) {
  crop_raster(p, possibilityLayer)
})

# Check output length
stopifnot(length(result_files) == length(plot_ids))

# Reset
plan(sequential)
```