---
title: "Sample Wet Points"
author: Maggie Church
created: "2024-09-24"
updated: "2025-02-24"
---

**Overview:** This script randomly samples 200 water points per plot-survey. In a way, sample design within a plot-survey is meant to stratify by pond. This outputs 735k water points, which I then upload to GEE to extract their predictors. 735k is way too many for training, but we wanted more than enough in case some were cloud-occluded.

**Inputs:**

-   Cleaned, in-buffered HAPET pond shapefile (script #3)

-   State boundary shapefile

**Steps:**

1. Build function to randomly sample a point within a geometry

2. Build function to randomly sample 1 point per pond, cycling through sampling if necessary to reach n points without replacement

3. Run function: get 200 water points per plot-survey

4. Add state attribute

**Output:**

-  A shapefile of water points, saved in `data/intermediate/sampled_coordinates/wet_200.shp`

**Note:** If I could go back in time and use "water" instead of "wet," I would. But I don't want to mess with that now. 

# Load libraries
```{r, message=F, warning=F, results='hide'}
library(tidyverse)
library(sf)
library(terra)
library(here)

utm14 <- 32614
wgs <- 4326

set.seed(123)
```

# Read data
Read in preprocessed, combined, and in-buffered HAPET data. Buffer allows 10-m pixels to be mostly wet. 
```{r}
# Load in-buffered HAPET data
allPonds_inbuff75 <- st_read(here("data", "intermediate", "buffered_ponds", "allPonds_inbuff75.shp")) 

# Load state shapfile
states <- st_read(here("data", "boundaries", "states"))
```

# 1. Function to safely sample 1 point from a geometry
```{r}
# Sample 1 point per pond with error handling
safe_sample <- function(geometry) {
  tryCatch({
    sampled_point <- st_sample(geometry, 1, force = TRUE)
    if (length(sampled_point) == 0) {
      return(NA)
    } else {
      return(sampled_point)
    }
  }, error = function(e) {
    return(NA)  # Return NA in case of error
  })
}
```

# 2. Function to sample n points without replacement
Randomly sample 1 point per geometry, cycling back through until n points are reached, ensuring no point is sampled more than once
```{r}
cycle_sample <- function(grouped_sf, n) {
  
  # Step 1: Sample 1 point per pond
  sampled_points <- grouped_sf |>
    mutate(sample_point = map(geometry, safe_sample)) |>
    filter(!map_lgl(sample_point, ~ all(is.na(.x))))  # Remove NAs
  
  n_ponds <- nrow(sampled_points)
  
  # Step 2: If there are fewer ponds than n, cycle through the ponds
  if (n_ponds < n) {
    remaining_needed <- n - n_ponds
    full_cycles <- floor(remaining_needed / n_ponds)
    leftover <- remaining_needed %% n_ponds
    
    # Repeat the full cycles
    full_sampled_points <- sampled_points[rep(1:n_ponds, full_cycles), ]
    
    # Add the remaining samples (without replacement, but ensuring no duplicates within this cycle)
    leftover_sampled_points <- sample_n(sampled_points, leftover, replace = FALSE)
    
    # Combine full cycles and leftover
    sampled_points <- bind_rows(sampled_points, full_sampled_points, leftover_sampled_points)
  }
  
  # Step 3: Ensure the final number of points is exactly `n`
  final_sample <- sampled_points |> slice_sample(n = n, replace = FALSE)
  
  return(final_sample)
}
```

# 3. Apply function
```{r}
# Apply the sampling function to all plots
wetPoints <- allPonds_inbuff75 |>
  group_by(Plot, dataset) |>
  group_modify(~ cycle_sample(.x, n=200)) |>
  ungroup() 

wetPoints_sf <- st_as_sf(
  data.frame(
    geometry = do.call(c, wetPoints$sample_point),  
    st_drop_geometry(wetPoints) 
  ) |> select(-sample_point) 
) |> st_set_crs(utm14) |> 
  distinct() 
```

# 4. Add state info
```{r}
# Get state info
states <- states |>
  st_transform(utm14) |> 
  select(STATEFP, STUSPS)

# Add state info to wet points
wetPoints_sf_wstate <- wetPoints_sf |>
  st_join(states) |>  # add state info
  st_transform(wgs)    # set to WGS before exporting
```

# Export
```{r}
st_write(wetPoints_sf_wstate, here("data", "intermediate", "sampled_coordinates", "wet_200.shp"), append=FALSE)
```