---
title: "Sample Dry Points"
author: Maggie Church
created: "2024-09-24"
updated: "2025-02-24"
---

**Description**: This script randomly samples 200 nonwater points per plot-survey. That is, we sample within the "possibility" layer where there's no known water (based on HAPET and JRC's GSW). This outputs 735k nonwater points, which I then upload to GEE to extract their predictors. 735k is way too many for training, but we wanted more than enough in case some were cloud-occluded.

**Inputs:**

-   Cleaned, out-buffered HAPET pond shapefile (script #3)

-   State boundary shapefile

-   Plot-sized permanent lake rasters (script #4)

-   Plot-sized "possibility" rasters (script #5)

**Steps:** For a given plot:

1.  Build function to sample nonwater points per plot-survey: load possibility raster for the plot --> mask edges of possible wetlands --> mask HAPET ponds for that plot-survey --> mask permanent water --> randomly sample from within the resulting dry ROI.

2.  Run the function: get 200 nonwater points per plot-survey

3. Add state attribute

**Outputs:**

-  A shapefile of water points, saved in `data/intermediate/sampled_coordinates/dry_200.shp`

**Note:** If I could go back in time and use "nonwater" instead of "dry," I would. But I don't want to mess with that now. 

## Load libraries
```{r, message=F, warning=F}
library(tidyverse)
library(sf)
library(terra)
library(future.apply)
library(parallel)
library(tools)
library(here)

w <- detectCores()

utm14 <- 32614
wgs <- 4326
```

## Load data
```{r, message=F, warning=F}
# Load out-buffered HAPET data
ponds_outbuff75 <- st_read(here("data", "intermediate", "buffered_ponds", "allPonds_outbuff75.shp"))

# Load state shapefile
states <- st_read(here("data", "boundaries", "states"))

# List raster files
cropped_poss_files <- list.files(here("data", "intermediate", "raster_processing", "cropped_poss_rasters"), full.names = TRUE) 
cropped_permWater_files <- list.files(here("data", "intermediate", "raster_processing", "cropped_permWater_rasters"), full.names = TRUE) 
```

## 1. Create fn to remove known wet areas from possibility layer, then sample dry points
```{r}
# Define a function to process each plot
sample_dry <- function(p, d) {
  
  # Skip plots 914 and 945 (these aren't within the PPR boundary)
  if (p %in% c(914, 945)) {
    message(paste("Skipping plot", p))
    return(NULL)  # Skip to the next plot
  }
  
  # 1. Get the possibility raster for this plot
  poss_file <- grep(paste0("_", p, "\\.tif$"), cropped_poss_files, value = TRUE)
  poss_i <- rast(poss_file)  # Read the file

  # 2. Only use non-edge pixels of possibility layer
  poss_i <- as.numeric(poss_i == 1)  
  poss_i <- mask(poss_i, poss_i, maskvalue = FALSE)  # Convert 0 to NA
  
  # 3. Remove actively wet areas from the possibility raster for this plot
  # Get the ponds for this plot
  ponds_i <- filter(ponds_outbuff75, dataset == d & Plot == p)
  
  # Get the permanent water for this plot
  perm_file <- grep(paste0("_", p, "\\.tif$"), cropped_permWater_files, value = TRUE)
  
  # Snap permanent water raster to poss_i -- add a 30m buffer too
  perm_i <- perm_file |>
      rast() |>        
      resample(poss_i) 
  
  # Convert 0 to NA
  perm_i <- mask(perm_i, perm_i, maskvalue = FALSE)   
  
  # add 30m buffer to the permanent water raster
  perm_i <- perm_i |> focal(w = matrix(1, 7, 7), fun = max, na.rm = TRUE) 
  
  # Add some areas that the perm water layer missed (I created this layer manually)
  perm_more <- st_read(here("data", "intermediate", "jrc_missed")) |> 
    st_transform(st_crs(perm_i)$wkt)

  # Mask water
  dryROI_subset <- poss_i |>
    mask(ponds_i, inverse = TRUE) |> 
    mask(perm_i, inverse = TRUE) |> 
    mask(perm_more, inverse = TRUE) 
  
  # 4. Sample points (200 per plot-year)
  sampled_points <- 
    spatSample(dryROI_subset, 
               size = 200, 
               method = "random", 
               na.rm = TRUE, 
               as.points = TRUE) |>
      st_as_sf()
  
  # If no points were sampled, return NULL
  if (nrow(sampled_points) == 0) {
    message(paste("No points sampled for plot", p))
    return(NULL)
  }

  # Add attributes
  sampled_points$plot_id <- p
  sampled_points$dataset <- d
  sampled_points$flyvr_d <- unique(pull(ponds_i, flyvr_d))
  
  return(sampled_points)
}
```

## 2. Run function
Btw plots 914 and 945 are missing from the possibility rasters because they're outside of the PPJV boundary.
```{r}
set.seed(123)

# set up parallel processing
plan(multisession, workers = w-2)

# Initialize a list to store results for all datasets
all_sampled_points_list <- list()

# get the dataset names
datasets <- ponds_outbuff75 |> st_drop_geometry() |> distinct(dataset) |> pull()

# Use lapply to apply the process to all plots, within a given dataset d
for (d in datasets) {

  # Plots surveyed in dataset d
  plots_d <- ponds_outbuff75 |>
    filter(dataset == d) |>
    pull(Plot) |>
    unique() |>
    na.omit() |>
    as.vector()
  
  # Sample dry points for this dataset
  sampled_points <- lapply(plots_d, sample_dry, d = d)
  sampled_points_sf <- bind_rows(sampled_points)
  
  # Append the dataset's sampled points to the main list
  all_sampled_points_list[[d]] <- sampled_points_sf
}

# combine the results into one sf object
dryPoints_sf <- bind_rows(all_sampled_points_list, .id = "dataset")

# Reset the plan to sequential
plan(sequential)
```

# 3. Add state info
```{r}
# Get state info
states <- states |>
  st_transform(utm14) |> 
  select(STATEFP, STUSPS)

# Add state info to points
dryPoints_sf_wstate <- dryPoints_sf |>
  st_join(states) |>  # add state info
  st_transform(wgs)    # set to WGS before exporting
```

```{r}
st_write(dryPoints_sf_wstate, here("data", "intermediate", "sampled_coordinates", "dry_200.shp"), append=FALSE)
```