---
title: "Select Training Data"
author: Maggie Church
created: "2024-08-16"
updated: "2025-02-24"
---

**Description**: This script determines training and testing plots... We selected these points in scripts 6 and 7, and extracted remotely-sensed predictors to them in GEE (code [here](https://code.earthengine.google.com/f84a065caa169012bab22b3b75319eae){.uri}).

**Inputs:**

-   Plot boundary shapefile (created in script #1)

-   PPJV administrative boundary

-   State boundary shapefile

-   Modified ecoregion boundaries shapefile (created in helper_functions/modify_ecoregions.R)

-   Shapefiles of point data with remotely-sensed features (see GEE link above)

**Steps:** 

1.  Modify Level-III ecoregions

2.  Select test surveys

3.  Select 1 cluster of plots within each ecoregion -- these are our test plots

4.  Cleanup the GEE export of water and nonwater data points with predictors (notably, drop occluded imagery)

5.  Remove test plots and test surveys from the GEE-exported dataset -- these are our candidate training points

6.  Add modified ecoregion to candidate training points (important attribute)

7.  Add CV blocking var to candidate training points (we'll use this for spatiotemporally aware splits in cross-validation)

8.  Add grid id to candidate training point (we'll use this for stratification, in selecting final training points)

9.  Create a dataset of test plot-surveys

**Outputs:**

-  csv of test plot-surveys

-  shapefile of candidate training points

-  shapefile of grid (just for our records)

## Load libraries
```{r}
library(tidyverse)
library(sf)
library(here)

set.seed(123)

source(here("code", "helper_functions", "VIs.R"))
source(here("code", "helper_functions", "perform_kmeans.R"))
```

# Load data
```{r}
# Load ppr boundary
ppr <- st_read(here("data", "boundaries", "PPJV")) |> st_transform(4326)

# read plot boundaries
plots <- st_read(here("data", "allPlots"))

# load modified ecoregion boundaries
ecoregions <- st_read(here("data/boundaries/ecoregions6/ecoregions6.shp"))

# Load state shapefile
states <- st_read(here("data", "boundaries", "states")) |> st_transform(4326)

################### 
# Define the folder path where shapefiles are located
folder_path_wet <- here("data", "gee_exports", "data_with_predictors", "wet_200_2wk")
folder_path_dry <- here("data", "gee_exports", "data_with_predictors", "dry_200_2wk")

# List all .shp files in the folder (ensure full path is returned)
shp_files_wet <- list.files(path = folder_path_wet, pattern = "\\.shp$", full.names = TRUE)
shp_files_dry <- list.files(path = folder_path_dry, pattern = "\\.shp$", full.names = TRUE)
shp_files <- c(shp_files_wet, shp_files_dry)

# Read each shapefile into a list of sf objects
shapefiles_list <- lapply(shp_files, function(file) {
  
  print(file)
  
  shp <- st_read(file, quiet = TRUE) |>
    mutate(type = ifelse(grepl("wet", basename(file)), "wet", "dry")) |> 
    st_set_crs(4326) 
  
  return(shp)
})
```

# 1. Pick test surveys
We selected August 2016, May 2022, and May 2024 as test surveys based on PHDI distributions
```{r}
testyrs <- c("brood16", "2022", "2024")
trainyrs_fullppr <- c("2019", "2021", "2023")
```

# 2. Pick test plots
Perform coordinate-based clustering of plots within each ecoregion.
The number of clusters (k) was chosen manually after visually exploring the spatial distribution of plots
Test clusters were then selected manually based on visual inspection
Outcome: Test plots comprise approximately 14% of all plots

Note: when I originally did this, I didn't set a seed :( so I can't perfectly recreate it. But this workflow represents what I did.
```{r}
# Cluster plots within ecoregions (5 clusters)
clusters5 <- plots |>
  st_centroid() |> 
  st_join(ecoregions, join=st_intersects) |>    #get the ecoregion for a plot's centroid
  group_by(L3mod) |> 
  group_split() |>
  lapply(perform_kmeans, k=5)                          

# Cluster plots within ecoregions (12 clusters - do this bc some ecoregions are too big)
clusters12 <- plots |>
  st_centroid() |> 
  st_join(ecoregions, join=st_intersects) |> 
  group_by(L3mod) |> 
  group_split() |>
  lapply(perform_kmeans, k=12) 

# Finally, do again for MT specifically (6 clusters)
clustersMT <- plots |>
  st_centroid() |> 
  st_join(ecoregions, join=st_intersects) |> 
  filter(L3mod=="Northwestern Glaciated Plains (MT)") |> 
  group_by(L3mod) |> 
  group_split() |>
  lapply(perform_kmeans, k=6) 

#################################################################
## select 1 test clusters for each ecoregion ## 
test_plot_ids_1 <- clusters5 |> 
  bind_rows() |> 
  filter((L3mod=="Western Corn Belt Plains" & cluster_id==5)
         | (L3mod=="North Central Hardwood Forests" & cluster_id==4)
         | (L3mod=="Lake Agassiz Plain" & cluster_id==4)) |>
  distinct(Plot) |>
  pull()

test_plot_ids_2 <- clusters12 |> 
  bind_rows() |>
  filter((L3mod=="Northern Glaciated Plains" & cluster_id==6)
         | (L3mod == "Northwestern Glaciated Plains (ND/SD)" & cluster_id==10)) |>
  distinct(Plot) |>
  pull()

test_plot_ids_3 <- clustersMT |> 
  bind_rows() |>
  filter(L3mod == "Northwestern Glaciated Plains (MT)" & cluster_id==1) |>
  distinct(Plot) |>
  pull()

# August 2016 (these are one-off bonus plots, so we gotta do this separately)
# Here's a manually-selected cluster of hold-out plots from August 2016: 
test_plots_aug16 <- c(997:1004)

# Combine test plot ids
test_plot_ids <- c(
  test_plot_ids_1,
  test_plot_ids_2,
  test_plot_ids_3,
  test_plots_aug16)

#################################################################
# make sure there's data from this plot-survey in all_samples (ie it's not totally clouded)
# get ponds 
test_plot_surveys_initial <- all_samples |> 
  st_drop_geometry() |>  
  distinct(plot_id, dataset) |> 
  filter(dataset %in% testyrs & plot_id %in% test_plot_ids) 
  
# don't trust HAPET on these
plot_drop_all  <- c(204, 620)
plot_drop_2022 <- c(271,824,414,202,803,827,804,746,264,269,312,313) 
plot_drop_2024 <- c(155, 267)

test_plot_surveys <- test_plot_surveys_initial |> 
  filter(!plot_id %in% plot_drop_all) |>
  filter(!(plot_id %in% plot_drop_2022 & dataset=="2022")) |>
  filter(!(plot_id %in% plot_drop_2024 & dataset=="2024")) 
```

# 3. Cleanup data points
Wrangling steps include: 
- I forgot to scale the S2 L1C values to 0-1, do that here 
- Drop if missing optical data (WE ONLY TRAIN ON UNOCCLUDED IMAGERY - that's because preliminary analysis showed a S1-only model wasn't great)
- Drop weird artifacts (band values > 1, NA/infinite VIs, etc)
- Calculate VIs
- Convert flyvr_d to date type
- Get a single plot_id col
```{r}
# vector of band_cols (we'll need to clean these bands a bit)
band_cols <- c("blue", "green", "red", "red_edge_1", "red_edge_2", "red_edge_3", "nir", "swir1", "swir2")

# Combine sf objects
all_samples <- do.call(bind_rows, shapefiles_list) |> 
  dplyr::select(-b1) |> 
  filter(!if_any(all_of(band_cols), ~ .x == -9999)) |>           # drop if missing S2 data
  mutate(across(all_of(band_cols),                               # rescale S2 L1C values
                ~ if_else(
                  dataset %in% c("pair16", "brood16", "pair17", "brood17"), 
                  .x * 10000, 
                  .x))
         ) |>
  filter(!if_any(all_of(band_cols), ~ .x < 0 | .x > 1)) |>       # drop values outside of 0 to 1
  filter(!(red==0 & nir==0)) |>                                  # drop this 1 weird row
  moreVIs() |>                                                   # calculate VIs
  filter(!is.na(NDRE))  |>                                       # drop this too 
  filter(!if_any(-geometry, is.infinite)) |>                     # drop rows with infinite values
  mutate(flyvr_d = as.Date(as.POSIXct(flyvr_d / 1000,            # convert flyvr_d to date
                                      origin = "1970-01-01",
                                      tz = "UTC"))) |>  
  mutate(plot_id = if_else(is.na(plot_id), Plot, plot_id)) |>    # keep only 1 plotid col
  select(-Plot) |> 
  filter(!is.na(plot_id))                                        # drop rows with missing plot ids (drops some dubious dry points)
```

# 4. Get training data
Get `all_samples` that are NOT within test years or test plots, for training
```{r}
training <- all_samples |> 
  filter(!(dataset %in% testyrs | plot_id %in% test_plot_ids)) 
```

# 5. Add modified ecoregion to training data
```{r}
training <- training |> st_join(ecoregions)  
```

# 6. Add CV blocking var to training data
Add blocking variable to training set - we'll use these for cross-validation in model selection
Use 2 clusters for full survey years (2019, 2021, 2023) and 1 cluster for the smaller survey years (2016-2017)
Ideally I'd use smaller spatial clusters, because our model doesn't need to be THAT generalizable, but then model selection takes forever
```{r}
training <- training |>
  filter(dataset %in% trainyrs_fullppr) |>
  group_split(dataset) |>
  lapply(perform_kmeans, k=2) |> 
  bind_rows() |> 
  bind_rows(training |> filter(!dataset %in% trainyrs_fullppr) |> mutate(cluster_id=1)) |> 
  ungroup() |> 
  mutate(block = paste(cluster_id, "-", dataset)) 
```

# 7. Add grid id to training data
Create a grid over the PPR (hexagonal, 51 cells), and add grid_id to the candidate training data. 
Then, drop grid-surveys with too few observations (we're going to stratify sampling on grid-surveys, and we don't want to oversample the heck out of these areas)
Note: this needed to be in a projected crs for better area calculation (we used utm14) - WGS created non-equal-area gridcells. Probably should've used something like Albers Equal-Area Conic, but it shouldn't make a difference in the end
```{r}
ppr_utm <- ppr |> st_transform(32614)
plots_utm <- plots |> st_transform(32614)

# create a hexagonal grid over the ppr
grid <- st_make_grid(ppr_utm, square=F, n=14) 
index <- which(lengths(st_intersects(grid, plots_utm)) > 0)
grid_ppr <- grid[index]

# convert the grid to sf
grid_sf <- grid_ppr |>  
  st_as_sf() |>
  mutate(grid_id = row_number()) 

# visualize grid
ggplot() + 
  geom_sf(data = ppr_utm, fill = "orange") + 
  geom_sf(data = grid_sf, fill = NA)

# calculate proportion overlap of each grid cell with the PPR (use a CRS that preserves area better than WGS84)
intersection <- st_intersection(grid_sf, ppr_utm) 
intersection$intersect_area <- st_area(intersection) 
intersection <- intersection |> mutate(grid_area_prop = intersect_area/max(intersect_area))
intersection <- intersection |> st_drop_geometry() |> select(grid_id, grid_area_prop)

# add grid_id to dataset and its proportion
training <- training |>  
  st_join(st_transform(grid_sf, 4326), join = st_intersects) |> # add grid_id
  left_join(intersection)                                       # add grid_area_prop

# remove grid-surveys with too few obs
training2 <- training |> 
  group_by(dataset, grid_id, type) |> 
  filter(n() > 30) |> 
  ungroup()
```

# 8. Get test plot-surveys
So for training points, we're pretty sure our wet and dry samples are correctly labeled (as much as HAPET's digitization are correct).
For testing, we'll use all pixels within plots. Instead of dealing with the sometimes-mislabeled permanent water, I just drop plots-surveys that seem suspect. 
I also further stratify the test set, picking 6 plots per ecoregion-survey
In the end, there are 128 test plot-surveys
```{r}
# get ponds 
test_plot_surveys_initial <- all_samples |> 
  st_drop_geometry() |>  
  distinct(plot_id, dataset) |> 
  filter(dataset %in% testyrs & plot_id %in% test_plot_ids) 
  
# don't trust HAPET on these
plot_drop_all  <- c(204, 620)
plot_drop_2022 <- c(271,824,414,202,803,827,804,746,264,269,312,313) 
plot_drop_2024 <- c(155, 267)

test_plot_surveys <- test_plot_surveys_initial |> 
  filter(!plot_id %in% plot_drop_all) |>
  filter(!(plot_id %in% plot_drop_2022 & dataset=="2022")) |>
  filter(!(plot_id %in% plot_drop_2024 & dataset=="2024")) 


# add ecoregion
ecoregion_plotid_xwalk <- plots |>
  st_centroid() |> 
  st_join(ppr_ecoregions_mod, join=st_intersects) |>
  distinct()

# Stratify plot- and iou-based datasets by time and space too. I pick 6 plots per ecoregion-dataset. For 2022 and 2024, I make sure # I select plots that are available for both years. 
test_plots_22_24 <-  all_plot_results_oost |>
  distinct(ecoregion, dataset, Plot) |>
  filter(dataset != "brood16") |>
  group_by(ecoregion, Plot) |>
  summarize(n=n()) |>
  filter(n == 2) |>
  ungroup() |>
  group_by(ecoregion) |>
  sample_n(6) |>
  ungroup() |>
  select(Plot)

test_plots_16 <- all_plot_results_oost |>
  distinct(ecoregion, dataset, Plot) |>
  filter(dataset == "brood16") |>
  distinct(Plot) |>
  sample_n(6)

test_plots <- bind_rows(test_plots_22_24, test_plots_16)
```

# Export training/testing data
```{r}
# save test plot-surveys, we'll use this again during model evaluation
write_csv(test_plot_surveys, here("data", "train_test_data", "test_plot_surveys.csv"))

# save grid
st_write(grid_sf, here("data", "train_test_data", "grid_sf", "grid_sf.shp"), append=F)

# save training dataset 
st_write(training, here("data", "train_test_data", "unbalanced_training", "training_2wk_200pt.shp"), append=F)
```