---
  title: "Determine Train/Test Split"
author: Maggie Church
date: "2024-08-16"
updated: "2025-02-24"
output: html_document
---
  This script splits dry and wet data points into training and testing. 

```{r}
train_yrs_fullppr <- c("2019", "2021", "2023")
testyrs <- c("brood16", "2022", "2024")
nd_trainyrs <- c("pair16", "pair17", "brood17")
```

```{r}
library(dplyr)
library(readr)
library(sf)
library(purrr)
library(here)

source(here("code", "helper_functions", "VIs.R"))
```

# read in GT points (2-wk composites)
Wrangling steps: 
  - rescale S2 L1C values
- drop values outside of -1 to 1
- convert flyvr_d to date
- calculate VIs
- get 1 nonnull plot_id col (drops some points, mostly dubious dry ones)

4wk - 200 wet points: raw nrow = 1,246,888
2wk - 200 wet points: raw nrow = 1,093,110
4wk - 1 point per pond: raw nrow = 
2wk - 1 point per pond: raw nrow = 

```{r}
set.seed(1)

# Define the folder path where shapefiles are located
folder_path_wet <- here("data", "gee_exports", "data_with_predictors", "PPR gtwPred-4wk", "wet_4wk_200")
folder_path_dry <- here("data", "gee_exports", "data_with_predictors", "PPR gtwPred-4wk", "dry_4wk")

# List all .shp files in the folder (ensure full path is returned)
shp_files_wet <- list.files(path = folder_path_wet, pattern = "\\.shp$", full.names = TRUE)
shp_files_dry <- list.files(path = folder_path_dry, pattern = "\\.shp$", full.names = TRUE)
shp_files <- c(shp_files_wet, shp_files_dry)

# Read each shapefile into a list of sf objects
shapefiles_list <- lapply(shp_files, function(file) {
  
  print(file)
  
  st_read(file, quiet = TRUE) %>%
    mutate(type = ifelse(grepl("wet", basename(file)), "wet", "dry")) %>% 
    st_set_crs(4326) %>% 
    return()
})

band_cols <- c("blue", "green", "red", "red_edge_1", "red_edge_2", "red_edge_3", "nir", "swir1", "swir2")

# Combine sf objects
all_samples <- do.call(bind_rows, shapefiles_list) %>% 
  dplyr::select(-b1) %>% 
  # rescale S2 L1C values
  mutate(across(all_of(band_cols),   
                ~ if_else(dataset %in% c("pair16", "brood16", "pair17", "brood17"), .x * 10000, .x))) %>%
  # drop values outside of -1 to 1
  filter(!if_any(all_of(band_cols), ~ .x < -1 | .x > 1)) %>% 
  # drop this 1 row
  filter(!(red==0 & nir==0)) %>% 
  # convert flyvr_d to date
  mutate(flyvr_d = as.Date(as.POSIXct(flyvr_d / 1000, origin = "1970-01-01", tz = "UTC"))) %>% 
  # keep only 1 plotid col
  mutate(plot_id = if_else(is.na(plot_id), Plot, plot_id)) %>% 
  select(-Plot) %>% 
  # Calculate VIs
  moreVIs() %>% 
  # drop rows with infinite values
  filter(!if_any(-geometry, is.infinite)) %>%
  # drop this too 
  filter(!is.na(NDRE)) 

# Some wet sampling occured outside of plot boundaries. Drop these. 
all_samples <- all_samples %>% filter(!is.na(plot_id)) 
```


For the 2wk data, we had valid pixels for 
40% of May 2016
46% of Aug 2016
90% of May 2017
22% of Aug 2017
69% of 2019
84% of 2021
92% of 2022
99% of 2023
92% of 2024

For the 4wk data, we had valid pixels for
67% of May 2016
95% of Aug 2016
100% of May 2017
90% of Aug 2017
90% of 2019
98% of 2021
99% of 2022
100% of 2023
99% of 2024

```{r, eval=F, echo=F}
# Calculate number of valid pixels (make sure you comment out the step in the chunk above where I filter values -1 to 1)
all_samples %>% 
  st_drop_geometry() %>%
  group_by(dataset) %>% 
  summarize(prop_valid = paste0(100-round(sum(red==-99990000 | red==-9999)/n()*100), "%"))
```

# Add modified Level-III ecoregion
```{r}
# load ecoregion boundaries
ecoregion5 <- st_read(here("data", "boundaries", "reg5_eco_l3")) %>% st_transform(4326)
ecoregion7 <- st_read(here("data", "boundaries", "reg7_eco_l3")) %>% st_transform(4326)
ecoregion8 <- st_read(here("data", "boundaries", "reg8_eco_l3")) %>% st_transform(4326)

# combine ecoregion layers
ecoregions <- bind_rows(ecoregion5, ecoregion7, ecoregion8) %>% 
  select(NA_L3NAME, NA_L2NAME)

# load ppr boundary
ppr <- st_read(here("data", "boundaries", "PPJV")) %>%
  st_transform(4326)

# get the ecoregions within the ppr
ppr_ecoregions <- st_intersection(ecoregions, ppr)

# modify ecoregion category, to consolidate (a region gets subsumed if <5 plots in it)
ppr_ecoregions2 <- ppr_ecoregions %>% 
  mutate(L3mod = case_when(
    NA_L3NAME %in% c("Northwestern Great Plains", "Middle Rockies", "Canadian Rockies", "Northwestern Glaciated Plains") ~ "Northwestern Plains",
    NA_L3NAME %in% c("North Central Hardwood Forests", "Northern Lakes and Forests") ~ "Northern Forests",
    NA_L3NAME %in% c("Lake Manitoba and Lake Agassiz Plain", "Northern Minnesota Wetlands") ~ "Lake Agassiz Plain",
    NA_L3NAME %in% c("Western Corn Belt Plains", "Driftless Area") ~ "Western Corn Belt Plains",
    T ~ NA_L3NAME
  ))

# add modified ecoregion to points
all_samples <- all_samples %>% st_join(ppr_ecoregions2) 
```

```{r}
# (re)equalize wet and dry points within a plot-survey 
all_samples <- all_samples%>%
  group_split(dataset, plot_id) %>%  # Split into list by dataset-plot
  map_dfr(~ {
    n_wet <- sum(.x$type == "wet")  # Count wet points in this plot
    sampled_dry <- .x %>%
      filter(type == "dry") %>%
      sample_n(size = min(n(), n_wet), replace = FALSE)  # Sample dry points
    bind_rows(filter(.x, type == "wet"), sampled_dry)  # Combine with all wet points
  })
```

# Determine testing set
I'll test on test on 2016 (wet) and 2019 (dry)
Randomly select plots for out-of-space testing set: 10% of plots within each ecoregion
```{r}
set.seed(1)

# sample 10% of plots from each L3mod, for full-region years 
test_plot_ids_fullyrs <- all_samples %>% 
  st_drop_geometry() %>% 
  # limit to wet points
  filter(type=="wet") %>%
  # limit to full-year training sets
  filter(!(dataset %in% testyrs) & !(dataset %in% nd_trainyrs)) %>% 
  # limit to surveys with at least 5 ponds
  group_by(plot_id, dataset) %>% 
  filter(n() > 5) %>% 
  ungroup() %>% 
  # sample 10% of plots from each ecoregion
  distinct(plot_id, L3mod) %>% 
  group_by(L3mod) %>% 
  sample_frac(0.1) %>% 
  distinct(plot_id) %>% 
  pull(plot_id) 

# sample 10% of plots for ND-only years
test_plot_ids_ndyrs <- all_samples %>% 
  st_drop_geometry() %>% 
  # limit to wet points
  filter(type=="wet") %>%
  # limit to nd-only training sets
  filter(dataset %in% nd_trainyrs) %>% 
  # limit to surveys with at least 5 ponds
  group_by(plot_id, dataset) %>% 
  filter(n() > 5) %>% 
  ungroup() %>% 
  # sample 10% of plots
  distinct(plot_id) %>% 
  sample_frac(0.1) %>% 
  pull(plot_id) 

# combine sampled plots
test_plot_ids <- c(test_plot_ids_fullyrs, test_plot_ids_ndyrs)
```

I want to compare a few methods (2 vs 4 wk, 200pt vs 1pt per pond)
and I want to make sure I'm comparing against the same test set.
So save this test set!
```{r}
write(test_plot_ids, here("data", "train_test_data", "test_plot_ids.txt"))
```

Read in test plot ids
```{r}
test_plot_ids <- scan(here("data", "train_test_data", "test_plot_ids.txt"))
```

```{r}
# get test from novel areas and years
testing <- all_samples %>% 
  filter(dataset %in% testyrs | plot_id %in% test_plot_ids) 

# get train data
training <- all_samples %>% 
  filter(!(dataset %in% testyrs | plot_id %in% test_plot_ids)) 
```

```{r}
set.seed(1)

# Function to perform k-means on each dataset's coordinates
perform_kmeans <- function(data) {
  coords <- sf::st_coordinates(data)  
  kmeans_result <- kmeans(coords, centers = 2, algorithm = "Lloyd", iter.max = 60)
  data$cluster_id <- kmeans_result$cluster
  return(data)  
}

# Split by dataset and apply kmeans to each group 
training <- training %>%
  filter(dataset %in% train_yrs_fullppr) %>%
  group_split(dataset) %>%
  lapply(perform_kmeans) %>% 
  bind_rows() %>% 
  bind_rows(training %>% filter(!dataset %in% train_yrs_fullppr) %>% mutate(cluster_id=1)) %>% 
  ungroup() %>% 
  mutate(block = paste(cluster_id, "-", dataset)) 
```

# Export training and testing
```{r}
st_write(training, here("data", "train_test_data", "unbalanced_training", "training_4wk_200pt.shp"), append=F)
write_csv(training, here("data", "train_test_data", "unbalanced_training", "training_2wk_200pt.csv"), append=F)
st_write(testing,  here("data", "train_test_data", "testing", "testing_4wk_200pt.shp"), append=F)
write_csv(testing,  here("data", "train_test_data", "testing", "testing_2wk_200pt.csv"))
```