---
title: "Training Curve Analysis"
author: "Maggie Church"
created: "2025-06-02"
updated: "2025-02-24"
---

**Description**: This script builds a training curve on the candidate training points. We do this to see how much training data we'll need for our model (we don't want too much or too little). 

**Inputs:**

-   PPJV administrative boundary

-   Candidate training data (created in script #8)

**Steps:** 

-   1. Prep the data for mlr3

-   2. Get learning curve data: train the model with sequentially more and more training data. Use the training sample design outlined in script #10, and spatiotemporally-aware cross-validation for model fitting/evaluation. 

**Outputs:**

-   Training curve results

**Note 1:** we don't tune the model features or hyperparameters.
**Note 2:** this takes a realllllyyyy long time and takes a looottt of memory. It fits 270 different models (10 x 9 x 3). Probably best run in an HPC. 

# Load libraries
```{r}
library(tidyverse)
library(sf)
library(here)
library(caret)
library(pROC)
library(future)
library(future.apply)
library(parallelly)
```

# Load data and helper function
```{r, message=F}
# load training data
training <- st_read(here("data", "train_test_data", "unbalanced_training", "training_2wk_200pt.shp")) |>
  rename(grid_area_prop = grd_r_p) #this got truncated in the export

# load PPR boundary
ppr <- st_read(here("data", "boundaries", "PPJV")) |> st_transform(32614)

# Load help function to perform spatiotemporal sampling
source(here("code", "helper_functions", "sptm_sample.R"))
```

# 2. Prep training data for training curve analysis
Create a vector of our candidate features, prep our training data for mlr3
```{r}
# full featureset (we'll use this later)
featureset <- c("VV","swir1","TCW","swir2","NDMI2","spei30d","NDRE", "NDVI",
                "sp30d_2","red","sp30d_1","nir","wt30_10","VVVH_rt",
                "wt90_10","green","z_2","NDWI","z_1","ABWI","h90_100","VH",
                "TCB","NDMI1","blue","h30_100","mNDWI2","mNDWI1","TCG",
                "rd_dg_1","z","rd_dg_2","rd_dg_3","NDPI","BU3","AWEI",
                "AWEISH","BSI","DVW","DVI","IFW","IPVI","MIFW","OSAVI","SAVI",
                "RVI","TVI","WRI","WTI", "VARI", "BGR", "GRR", "RBR", "SRBI", "NPCRI",
                "EVI3b", "EVI2b")  

# prep training data
training <- training |>
  st_drop_geometry() |> 
  mutate(type = as.factor(type)) |>
  mutate(grid_area_prop = as.numeric(grid_area_prop))
```

# 3. Get learning curve data 
Run this in parallel:
 *  Sample (stratified) from candidate training data, taking increasingly higher proportions of data from the full training set: 1%, 6%, 11% ... up to 46%
 *  For each proportion slice, fit an RF model on 8 of the 9 spatiotemporally-aware folds (created in script #8), optimizing for AUC
 *  Evaluate test AUC on held-out fold 
 *  Redo this 3 times per proportion slice-fold, selecting different training data each time
```{r, eval=F}
# set up parallel processing (careful, this is quite memory-intensive)
w <- parallelly::availableCores()
future::plan("multisession", workers=2)

# Grid of lappy parameters
props <- seq(.01, 0.46, by=0.05)  # 10 proportions (start with 1% of data, 6%, 11%, ... up to 46%)
folds <- unique(training$block)   # 9 folds
reps <- 1:3                       # 3 repetitions
params <- expand.grid(f = folds, p = props, r = reps) 

# run in parallel
results <- future_lapply(1:nrow(params), function(idx) {
  
  f <- params$f[idx]        # fold (for train-test split)
  p <- params$p[idx]        # proportion
  r <- params$r[idx]        # repetition

  tryCatch({
    
    # Set a new seed for each fold
    set.seed(1000 + r * 10 + round(p * 10000)) 
    
    # Split by block 
    test_data  <- training |> filter(block == f)    # set aside 1 fold for test
    train_pool <- training |> filter(block != f)    # sample from the other 8

    # Get number of rows in training data
    n_train_pool <- nrow(train_pool)

    # Get number of grid-data strata, weighted by grid_area_prop
    weighted_nstrata <- train_pool |>
      distinct(grid_id, dataset, grid_area_prop) |>
      summarize(sum(grid_area_prop)) |> 
      pull()

    # Get n. We'll sample n x p per strata
    n <- floor(n_train_pool / weighted_nstrata * p)

    # Sampling from train_pool data (stratify by class, grid_id, and survey)
    sample <- sptm_sample(train_pool, n_per_cell = n) 

    # Prep sample for ml3
    train_data <- sample |> select(all_of(featureset), type)

    # Each class could be present within each grid-survey sample, but just in case: 
    if (length(unique(train_data$type)) < 2 || length(unique(test_data$type)) < 2) {
      warning(sprintf("Skipping fold %d, prop %.3f: only one class present", f, p))
      return(tibble(fold = f, proportion = p, training_n = nrow(train_data), ROC_AUC = NA_real_))
    }

    # train a model on the training sample
    model <- train(
      type ~ .,
      data = train_data,
      method = "rf",
      metric = "ROC",
      trControl = trainControl(
        method="none",
        classProbs = TRUE,
        summaryFunction = twoClassSummary
      )
    )

    # apply the model to the test set
    probs <- predict(model, test_data, type = "prob")[, 2]

    # get test accuracy
    roc_auc <- roc(response = test_data$type, predictor = probs)$auc

    # return results
    tibble(fold = f, 
           rep = r,
           proportion = p, 
           training_n = n_train_pool,
           ROC_AUC = as.numeric(roc_auc))

  }, 
  
  # if this run fails for some reason, print a warning and move on
  error = function(e) {
    warning(sprintf("fold %d, prop %.3f failed: %s", f, p, e$message))
    return(tibble(fold = f, proportion = p, training_n = NA_integer_, ROC_AUC = NA_real_))
  })

}, future.seed = TRUE) |> bind_rows()

# turn off parallel processing
future::plan("sequential")
```


```{r}
summary_results <- results |>
  group_by(proportion) |>
  summarise(
    mean_n = mean(training_n, na.rm=T),
    mean_auc = mean(ROC_AUC, na.rm = TRUE),
    sd_auc = sd(ROC_AUC, na.rm = TRUE),
    se_auc = sd_auc / sqrt(n()),
    lower_ci = mean_auc - 1.96 * se_auc,
    upper_ci = mean_auc + 1.96 * se_auc,
    .groups = "drop"
  )

ggplot(summary_results, aes(x = mean_n, y = mean_auc)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.2, fill = "blue") +
  geom_point(color = "blue", size = 2) +
  labs(
    title = "Learning Curve",
    subtitle = "Mean test AUC across 10 resamples. Dataset was spatiotemporally stratified, then split into 80-20 train/test splits.",
    x = "Training Data Size",
    y = "Mean AUC (with 95% CI)"
  ) +
  theme_minimal(base_size = 14)
```

Ok performance seems to top-out at the 41% of training mark. How many samples is that per cell-survey? ==> ~750
```{r}
floor(nrow_training / ngroups * 0.41)
```

```{r}
# save training curve results
write_csv(results, here("data/learning_curve_results/learning_curve_results_updated.csv"))
```
