---
title: "Balance Training Data"
author: "Maggie Church"
created: "2024-10-12"
updated: "2025-02-24"
---

**Description**: This script selects final training points from the candidate training data by stratifying over time and space. 

**Inputs:**

-   PPJV administrative boundary

-   Candidate training data (created in script #8)

**Steps:** 

-   Balance training data spatiotemporally: draw 750 water and 750 nonwater points, from each grid-survey (750 was informed by the training curve from script #9)

**Outputs:**

-   Final training dataset

# Load libraries
```{r, message=F}
library(tidyverse)
library(sf)
library(here)

set.seed(123)
```

# Load data and helper function
```{r, message=F}
# Load PPR boundary
ppr <- st_read(here("data", "boundaries", "PPJV")) |> 
  st_transform(32614)

# Load candidate training data
training <- st_read(here("data", "train_test_data", "unbalanced_training", "training_2wk_200pt.shp")) |> 
  st_transform(32614)

# Load help function to perform spatiotemporal sampling
source(here("code", "helper_functions", "sptm_sample.R"))
```


## Balance dataset
My goal is to balance samples spatially across the "grid," and temporally within each discrete survey period. Basically, I don't want to over-train in one region/period. Remaining data gaps include:
-   W. MT has none 2019
-   2016-2017 is ND-only
```{r}
# sample each cell-survey: weight by spatial coverage of that survey (see dataset_weights df), and proportion of much of the gridcell is inside the ppr. Ensure even wet-dry sampling
balanced_training <- sptm_sample(training, n_per_cell = 750) |>
  # remove cells where there were no wet points
  group_by(dataset, grid_id) |> 
  filter(!is.na(sum(type=="wet")) & sum(type=="wet") > 0) |> 
  ungroup() 

# some light cleanup
balanced_training_clean <- balanced_training |>
  # remove cells where there were no wet points
  group_by(dataset, grid_id) |> 
  filter(!is.na(sum(type=="wet")) & sum(type=="wet") > 0) |> 
  ungroup()  |>
  # remove these sneaks, too
  filter(!is.na(NDRE) & !is.na(NDMI2))
```

# Export
Write balanced training dataset to  csv, because it's way easier to deal with in the next scripts (which I run on an HPC)
```{r}
write_csv(balanced_training_clean, here("data", "train_test_data", "balanced_training", "training_2wk_200pt_bal.csv"))
```
