---
title: "2a. Process Possibility Rasters"
output: html_notebook
date: "2024-09-24"
author: "maggie church"
---

```{r}
library(sf)
library(dplyr)
library(terra)
library(future.apply)
library(parallel)
library(tools)

utm14 <- 32614
w <- detectCores()
temp_dir <- "../data/intermediate" 
raw_dir <- ("~/Documents/PPR/raw_data/GreatPlains/raw_data/")  # raw HAPET data folder
```

**Steps:**

1.  Reproject and self-mask possibility raster tiles. Save output to intermediate data folder.

2.  Break the larger tiles into plot-size rasters. Save all 651 plot-size rasters to intermediate data folder.

**Inputs**:

-   Plot boundaries (note these don't apply to the 2016-2017 data).

-   Possibility layer tiles. I created the Possibility Layer in GEE, [here](https://code.earthengine.google.com/388a0678b7a0ff8d0705dbdaccf47e4c){.uri}. Raster format, epsg:4326 crs.

### 1. reproject and self-mask the larger tiles

I exported the possibility layer from GEE, which split the region into 8 tiles. I should have exported straight to UTM Zone 14N, but instead I just deal with that reprojection here. I also self-mask the rasters here. Save to intermediate data folder. 

This first step takes a few hours :/
```{r}
# setup parallel backend (uses all available cores by default)
plan(multisession, workers=w-2) 

# read in the list of possibility layer files
files <-  list.files('../data/inputs/gee_rasters/possibilityLayer', full.names = TRUE)

# Parallelize the raster loading, projecting, and masking
loaded_poss_files <- future_lapply(files, function(file) {
  
  # Load and reproject the raster
  rast_proj <- rast(file) %>% project(utm14_str, res = 10)
  
  # Mask out 0s (or FALSE values) in the raster
  masked_rast <- mask(rast_proj, rast_proj, maskvalue = FALSE)
  
  # Write to disc...
  temp_file <- paste0(temp_dir, "rastMask", file_path_sans_ext(basename(file)), ".tif")
  writeRaster(masked_rast, temp_file)
  
  # but also return here...
  return(masked_rast)
})

# Reset the plan to sequential (no parallelization)
plan(sequential)
```

### 2. Break (processed) possibility raster tiles into plot-size rasters

Load plot boundaries and apply buffer.
```{r, message=F, warning=F}
# read in plot boundary data
plots1617 <- 
  st_read("../data/intermediate/plots1617/plots_1617.shp") %>%
  st_transform(utm14) %>%
  rename(Plot=plot_id)

plots <- 
  st_read(dsn = paste0(raw_dir, 'FWS_2018_2019/FSMS_Enterprise_Albers.gdb'), layer='FSMS_PLOTS') %>%
  st_transform(utm14) %>% 
  bind_rows(plots1617)

# apply inwards plot buffer 
plots_inbuff75 <- st_buffer(plots, -7.5)
```

This next chunk creates the function we'll use to split these Possibility tiles up. If a plot covers multiple tiles, crops from each tile are merged to cover full plot extent.
```{r}
# Function to crop raster tiles by plots
crop_raster <- function(p, rasters) {
  
  # Get the extent for the current plot by filtering the plot_id
  plot_extent <- plots_inbuff75[plots_inbuff75$Plot == p, ]
  
  # Initialize a list to store cropped rasters
  cropped_rasters <- list()

  # Initialize a list to store cropped rasters
  for (i in seq_along(rasters)) {
    raster <- rasters[[i]]  # Extract the raster from the list

        # Crop the raster with error handling for non-overlapping extents
    cropped_raster <- tryCatch({
        crop(raster, plot_extent)
      }, error = function(e) {
        return(NULL)})
    
  # If a valid cropped raster was returned, store it in the list
  if (!is.null(cropped_raster) && !all(is.na(values(cropped_raster)))) {
    cropped_rasters <- c(cropped_rasters, list(cropped_raster))
    }
  }
  
  # If there are any cropped rasters, merge them or return the single raster
  if (length(cropped_rasters) > 0) {
    if (length(cropped_rasters) == 1) {
      # If there's only one cropped raster, no need to merge, just return it
      merged_raster <- cropped_rasters[[1]]
    } else {
      # Merge all cropped rasters if more than one
      merged_raster <- do.call(terra::merge, cropped_rasters)
    }
    
    # Define the output file name based on the plot ID
    output_file <- paste0(temp_dir, "/cropped_poss_rasters/raster_", p, ".tif")
    
    # Save the merged raster
    writeRaster(merged_raster, output_file, overwrite = TRUE)
    
    return(output_file)  # Return the file path of the cropped raster
  } else {
    return(NULL)  # Return NULL if no rasters were cropped
  }
}
```

Now run the function! I use parallel processing to speed things up... this part only takes a few minutes. This write the plot-sized possibility rasters to the intermediate data folder. 
```{r}
# Get unique plot IDs
plot_ids <- unique(plots_inbuff75$Plot)

# set up parallel processing
plan(multisession, workers = w-2)

# Load all raster tiles from step 2
files_processed <- list.files(paste0(temp_dir, "/possibility_utm14N"), full.names = TRUE)
rasters <- lapply(files_processed, rast)

# Apply the function to each plot
result_files <- lapply(plot_ids, function(p) {
  crop_raster(p, rasters)
})

# check that output = number of plots
stopifnot(length(result_files) == length(plot_ids))

# Reset the plan to sequential (no parallelization)
plan(sequential)
```
