---
title: "Process Permanent Lakes"
author: Maggie Church
date: "2024-09-24"
updated: "2025-02-24"
output: html_document
---

This script reads in the Permanent Lake layer (created in GEE, consists of 6 large tiles) and crops it into 651 rasters that overlay each of our plots.

Link to GEE code that creates the Permanent Lake layer [here](https://code.earthengine.google.com/967bd27fac16af729ef93dc1fd53d1e0){.uri}. Note this is EPSG=32614 (UTM 14N)

# Setup environment and load libraries
```{r}
library(sf)
library(dplyr)
library(terra)
library(future.apply)
library(parallel)
library(tools)
library(here)

utm14 <- 32614
w <- detectCores()
```

# 1. Break permanent lake raster tiles into plot-size rasters

Load plot boundaries and apply buffer.
```{r}
# read in plot boundary data
plots <- 
  st_read(here("data", "allPlots")) %>%
  st_transform(utm14)

# apply inwards plot buffer 
plots_inbuff75 <- st_buffer(plots, -7.5)
```

This next chunk creates the function we'll use to split these tiles up. If a plot covers multiple tiles, crops from each tile are merged to cover full plot extent.

p is a plot id
rasters is a list of large rasters
```{r}
# Function to crop raster tiles by plots
crop_raster <- function(p, rasters) {
  
  # Get the extent for the current plot by filtering the plot_id
  plot_extent <- plots_inbuff75[plots_inbuff75$Plot == p, ]
  
  # Initialize a list to store cropped rasters
  cropped_rasters <- list()

  # Initialize a list to store cropped rasters
  for (i in seq_along(rasters)) {
    raster <- rasters[[i]]  # Extract the raster from the list
    
    # Crop the raster with error handling for non-overlapping extents
    cropped_raster <- tryCatch({
        crop(raster, plot_extent)
      }, error = function(e) {
        return(NULL)})
    
  # If a valid cropped raster was returned, store it in the list
  if (!is.null(cropped_raster) && !all(is.na(values(cropped_raster)))) {
    cropped_rasters <- c(cropped_rasters, list(cropped_raster))
    }
  }
  
  # If there are any cropped rasters, merge them or return the single raster
  if (length(cropped_rasters) > 0) {
    if (length(cropped_rasters) == 1) {
      # If there's only one cropped raster, no need to merge, just return it
      merged_raster <- cropped_rasters[[1]]
    } else {
      # Merge all cropped rasters if more than one
      merged_raster <- do.call(terra::merge, cropped_rasters)
    }
    
    # Define the output file name based on the plot ID
    output_file <- paste0(here("data", "intermediate", "raster_processing", "cropped_permWater_rasters"), "/raster_", p, ".tif")
    
    # Save the merged raster
    writeRaster(merged_raster, output_file, overwrite = TRUE)
    
    return(output_file)  # Return the file path of the cropped raster
  } else {
    return(NULL)  # Return NULL if no rasters were cropped
  }
}
```

Now we run the function! I use parallel processing to speed things up... this part only takes a few minutes.
```{r}
# Get unique plot IDs
plot_ids <- unique(plots_inbuff75$Plot)

# set up parallel processing
plan(multisession, workers = w-2)

# Load all 6 big raster tiles 
files <- list.files(here("data", "gee_exports", "gee_rasters", "permanentWater"), full.names = TRUE)
rasters <- lapply(files, rast)

# Apply the function to each plot
result_files <- lapply(plot_ids, function(p) {
  crop_raster(p, rasters)
})

# Reset the plan to sequential (stop parallelization)
plan(sequential)
```
