---
title: "R Notebook"
output: html_notebook
---

```{r, message=F, warning=F, results='hide'}
library(sf)
library(dplyr)
library(mapview)
library(purrr)
library(tidyr)
library(stringr)
library(terra)

# set directory where raw data lives
raw_dir <- ("~/Documents/PPR/raw_data/GreatPlains/raw_data/")

utm14 <- 32614
wgs <- 4326
```

Read in preprocessed and combined HAPET data (this is the output of `1.combineRawHapet.Rmd`).

```{r}
# Load HAPET data
allPonds <- st_read("../data/allPonds/allPond.shp")
```

Apply a 7.5m inward buffer to ponds, to ensure 10-m pixels are mostly wet

```{r}
# Remove pond edges
allPonds_inbuff75 <- 
  st_buffer(allPonds, -7.5) %>% 
  filter(st_is_empty(geometry)==FALSE)  # drop ponds that get "buffered out"
```

Sample wet GT points: 1 point from each pond.

```{r}
set.seed(123)

# Function to sample 1 point from each pond
safe_sample <- function(geometry) {
  tryCatch({
    sampled_point <- st_sample(geometry, 1, force = TRUE)
    if (length(sampled_point) == 0) {
      return(NA)
    } else {
      return(sampled_point)
    }
  }, error = function(e) {
    return(NA)  # Return NA in case of error
  })
}

# Sample points with error handling
 wetPoints <- allPonds_inbuff75 %>%
   mutate(sample_point = map(geometry, safe_sample)) %>%
   filter(!map_lgl(sample_point, ~ all(is.na(.x))))  

# Change the geometry from pond polygons to sample points
 wetPoints_sf <- st_as_sf(
  data.frame(
    geometry = do.call(c, wetPoints$sample_point),
    st_drop_geometry(wetPoints)
  ) %>% select(-sample_point)
 ) %>% st_set_crs(utm14)
```

Export wet points

```{r}
# Get state info
states <- st_read('./../../states') %>%
  st_transform(utm14) %>% 
  select(STATEFP, STUSPS)

# Add state info to wet points
wetPoints_sf_wstate <- wetPoints_sf %>%
  st_join(states) %>%  # add state info
  st_transform(wgs)    # set to WGS before exporting

st_write(wetPoints_sf_wstate, "../data/sampled_points/wet.shp", append=FALSE)
#wet_points <- st_read("../data/sampled_points/wet.shp")
```

Next, I upload the exported wet points to GEE to extract their predictors, then save those results at PPR/samplePoints/data/inputs.

Code for further analysis is found at `PPR/samplePoints/code`.
