---
title: "sample_gtwpred_data"
output: pdf_document
date: "2024-08-16"
---

```{r}
library(dplyr)
library(readr)
library(sf)

set.seed(123)
```


Wet points were sampled from raw HAPET data in PPR/raw/GreatPlains/wrangle_raw_maggie/samplePoints.Rmd
From there, the wet points are uploaded to GEE to extract predictor values.
Dry points are also sampled in GEE.
Wet and dry points with predictor values are saved in PPR/samplePoints/data/inputs

This script samples from all points, returning a subset of data for training and testing

# read in dry GT points
```{r}
dry16_17 <- st_read('../data/inputs/dry_wpred/dry1617/dry1617.shp')

dry19 <- st_read('../data/inputs/dry_wpred/dry19/compDry19.shp')

dry20 <- st_read('../data/inputs/dry_wpred/dry20/compDry20.shp')

dry21 <- st_read('../data/inputs/dry_wpred/dry21/compDry21.shp')

dry22 <- st_read('../data/inputs/dry_wpred/dry22/compDry22.shp') 

allDry <- bind_rows(dry16_17, dry19, dry20, dry21, dry22) %>% 
  filter(red_edge_3 != -9999 ) %>% # drop obs without optical data
  mutate(type = 'dry') %>%
  dplyr::select(-b1) %>% 
  mutate(flyvr_d = as.Date(as.POSIXct(flyvr_d / 1000, origin = "1970-01-01", tz = "UTC")))
```

# read in wet GT points
I should probably drop obs without optical data earlier, before sampling
```{r}
wet16_17 <- st_read('../data/inputs/wet_wpred/wet_1617/wet2.shp') 

wet19 <- st_read('../data/inputs/wet_wpred/wet19/compWet19.shp') 

wet20 <- st_read('../data/inputs/wet_wpred/wet20/compWet20.shp')

wet21 <- st_read('../data/inputs/wet_wpred/wet21/compWet21.shp')

wet22 <- st_read('../data/inputs/wet_wpred/wet22/compWet22.shp') 

allWet <- bind_rows(wet16_17, wet19, wet20, wet21, wet22) %>%
  mutate(type = 'wet') %>%
  filter(red_edge_3 != -9999) %>% # drop obs without optical data 
  dplyr::select(-b1) %>%
  mutate(flyvr_d = as.Date(as.POSIXct(flyvr_d / 1000, origin = "1970-01-01", tz = "UTC")))
```

Looks like we found valid pixels for >99% of points in all years (96-99% in 2017)
CHECK THIS
```{r}
# put all the training data together
all_samples <- bind_rows(allWet, allDry) 

all_samples %>% 
  st_drop_geometry() %>%
  group_by(type, dataset, STUSPS, sz_clss) %>% 
  summarize(n()) %>%
  arrange(STUSPS, type, sz_clss, dataset) %>%
  write_csv("../data/inputs/data_summary.csv")

# add more VIs
moreVIs <- function(df){ 
  df %>% 
    mutate(
      NDPI = (swir1-green)/(swir1+green),
      BU3 = red + swir1 - nir,
      AWEI = (green-swir2) - (0.25*nir + 2.75*swir1),
      AWEISH = blue + 2.5*green - 1.5*(nir+swir1) - 0.25*swir2,
      BSI = ((swir1+red-nir-blue)/(swir1+red+nir+blue))*100 + 100,
      DVW = NDVI-NDWI,
      DVI = nir-red,
      IFW = nir-green,
      IPVI = nir/(nir+red),
      MIFW = swir1-green,
      OSAVI = (nir-red)/(nir+red+0.16),
      SAVI = ((nir-red)/(nir+red+0.5))*1.5,
      RVI = nir/red,
      TVI = 0.5*(120*(nir-green)-200*(red-green)),
      WRI = (green+red)/(nir+swir1),
      WTI = 0.91*red + 0.43*nir,
      VARI = (green-red)/(green+red-blue),
      BGR = blue/green,
      GRR = green/red,
      RBR = red/blue,
      SRBI = blue-red,
      NPCRI = (red-blue)/(red+blue), # Normalized Pigment Chlorophyll Ratio Index 
      EVI3b = 2.5 * (nir - red)/(nir + (6*red) - (7.5*blue) + 1),
      EVI2b = 2.5 * (nir - red)/ (nir + (2.4*red) + 1)
    )
}

all_samples <- moreVIs(all_samples)

all_samples %>%
  st_write("../data/inputs/all_samples/all_samples.shp", append=F)
```

get test regions
```{r}
all_samples <- st_read("../data/inputs/all_samples/all_samples.shp") %>%
  filter(!is.na(dataset))

# load ecoregion boundaries
ecoregion5 <- st_read("../../ecoregions/reg5_eco_l3.shp") %>% st_transform(4326)
ecoregion7 <- st_read("../../ecoregions/reg7_eco_l3.shp") %>% st_transform(4326)
ecoregion8 <- st_read("../../ecoregions/reg8_eco_l3.shp") %>% st_transform(4326)

# combine ecoregion layers
ecoregions <- bind_rows(ecoregion5, ecoregion7, ecoregion8) %>% 
  select(NA_L3NAME, NA_L2NAME)

# load ppr boundary
ppr <- st_read("../../ppr_boundary/PPJV/PPJV.shp") %>% st_transform(4326)

# get the ecoregions within the ppr
ppr_ecoregions <- st_intersection(ecoregions, ppr)

# modify ecoregion category, to consolidate
ppr_ecoregions2 <- ppr_ecoregions %>% 
  mutate(L3mod = case_when(
    NA_L3NAME %in% c("Northwestern Great Plains", "Middle Rockies", "Canadian Rockies", "Northwestern Glaciated Plains") ~ "Northwestern Plains",
    NA_L3NAME %in% c("North Central Hardwood Forests", "Northern Lakes and Forests") ~ "Northern Forests",
    NA_L3NAME %in% c("Lake Manitoba and Lake Agassiz Plain", "Northern Minnesota Wetlands") ~ "Lake Agassiz Plain",
    NA_L3NAME %in% c("Western Corn Belt Plains", "Driftless Area") ~ "Western Corn Belt Plains",
    T ~ NA_L3NAME
  ))

# add modified ecoregion to points
all_samples <- st_join(all_samples, ppr_ecoregions2)

# table of wet points by modified ecoregions
all_samples %>% st_drop_geometry() %>% filter(type=="wet") %>% 
  group_by(NA_L3NAME, dataset) %>% summarize(n()) %>% gt::gt()

```

# Split training/testing
I'll test on test on 2016 (wet) and 2021 (dry)

Here, explore gridded data by ecoregion to select test regions
```{r}
set.seed(123)

testyrs <- c("pair16", "brood16", "2021")

all_samples_wet_trainyears <- all_samples %>% 
  filter(type=="wet" & !(dataset %in% testyrs))

# add a grid
grid <- st_make_grid(ppr, n=25) 
index <- which(lengths(st_intersects(grid, ppr_ecoregions2)) > 0)
grid_ppr <- grid[index]

# convert the grid to sf
grid_sf <- grid_ppr %>%  
  st_as_sf() %>%
  mutate(grid_id = row_number()) 

# actually let's save this, for CV partitioning later
# st_write(grid_sf, '../data/inputs/grid/grid_sf.shp', append=F)
```


Select test "boxes

I could do st_make_grid(n=50) above, then even test sets would have 150-180 GT points.
Test grid_ids for those are c(247, 118, 38, 174, 255)
```{r}
library(ggplot2)
library(mapview)


# find contenders for testing boxes
contenders <- grid_sf %>%
  st_join(all_samples_wet_trainyears, join = st_intersects) %>%
  group_by(grid_id) %>% 
  # the box falls into exactly 1 ecoregion
  filter(n_distinct(L3mod) == 1) %>%    
  # the box has samples from every dataset (ND will have more, from 2017) OR the box is in MT
  filter(case_when(
    L3mod == c("Northwestern Plains") ~ n_distinct(dataset) == 5,
    TRUE ~ n_distinct(dataset) == 3
  ) | STUSPS == "MT") %>% 
  ungroup() %>%
  group_by(grid_id, L3mod, dataset) %>% 
  summarize(n=n())                    

# I check these out in mapview and chose test boxes
# (I actually chose a site in MT that doesn't have data from 2020)
test_grid_ids <- c(120, 260, 201, 170, 209, 37)

mapview(grid_sf %>% filter(grid_id %in% test_grid_ids), col.regions="navy", legend=F) +
  mapview(ppr_ecoregions2, zcol="L3mod")

# limit to selected test boxes
test_boxes <- grid_sf %>%
  filter(grid_id %in% test_grid_ids)
```

There might have been an issue with export of some fields: Shape_Area, flyvr_d, Acres...
```{r}
# get test from novel areas and years
testing <- all_samples %>% 
  filter(dataset %in% testyrs) %>%
  bind_rows(st_filter(all_samples, test_boxes)) %>%
  distinct() %>% 
  st_join(grid_sf, join = st_intersects) # add grid_id
 
# get train data
training <- all_samples %>% 
  filter(!(dataset %in% testyrs)) %>%
  st_difference(st_union(test_boxes)) %>% 
  st_join(grid_sf, join = st_intersects) # add grid_id
```

# Export training/testing data
```{r}
st_write(training, '../data/intermediate/unbalanced_training/training.shp', append=F)
st_write(testing, '../data/intermediate/testing/testing.shp', append=F)
```
