---
title: "Determine decision Threshold"
author: "Maggie Church"
created: "2025-07-14"
updated: "2026-03-23"
---

**Description**: This script takes the confusion matrices (TN, TP, FN, TP) at different decision thresholds, using the final model features (script #11) and hyperparameters (script #12). These metrics are based on all pixels (within wetland footprints) from all training plot-surveys. I used all pixels, rather than the sampled training dataset, because I wanted to determine the threshold based on uneven classes (i.e. there are far more dry pixels than wet). And because that's a TON of data, I found this GEE-R process most efficient. 

**Inputs:**

-  Confusion matrix at different decision thresholds for all pixels in training test-surveys (exported from GEE, see link below)
-  Confusion matrix at different decision thresholds for all pixels in training test-surveys (exported from GEE), using Dynamic World
-  PPJV administrative boundary shapefile 
-  Plot boundaries shapefile (created in script #1)
-  Modified ecoregion boundaries shapefile (created in helper_functions/modify_ecoregions.R)

**Steps:** 

-  Create a function to calculate F-beta score at different thresholds
-  Find the optimal decision threshold for our model and dynamic world, where beta = 0.5
-  Explore how well this threshold serves different ecoregions (do we need ecoregion-specific threshold instead?)

**Outputs:**

-   A number: the decision threshold to use for classification (one for our model and one for dynamic world)



```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(sf)
library(pROC)
library(scales)
library(cowplot)
library(future)
library(furrr)
```

# Load data

Load plots, ppr, and ecoregion boundaries
```{r}
plots <- st_read(here("data/allPlots")) |> st_transform(32614)
ppr <- st_read(here("data", "boundaries", "PPJV")) |> st_transform(32614)
ecoregions <- st_read(here("data", "boundaries", "ecoregions6")) |> st_transform(32614)
```

# Create function to calculate f-beta score
Function to calculate f-scores for each threshold (using the df output from the GEE code), a specified beta
```{r}
get_f_by_thresh <- function(df, fbeta){
  df |> 
    group_by(threshold) |>
    summarize(
      TP = sum(TP),
      TN = sum(TN),
      FP = sum(FP),
      FN = sum(FN)
    ) |> 
    mutate(
      precision = TP/(TP+FP),
      recall = TP/(TP+FN),
      f = (1 + fbeta^2) * precision * recall / (fbeta^2 * precision + recall)
    )
}
```

# Optimal threshold for our model
OLD Load in our model's training probabilities 
```{r}
# point_probabilities_train <-
#   bind_rows(
#     read_csv(here("data/accuracy_assessment/ver3/point_probabilities_train/point_probabilities_May23_train.csv")),
#     read_csv(here("data/accuracy_assessment/ver3/point_probabilities_train/point_probabilities_May21_train.csv")),
#     read_csv(here("data/accuracy_assessment/ver3/point_probabilities_train/point_probabilities_May19_train.csv"))
#   ) |> 
#   mutate(dataset=as.character(dataset)) |>
#   bind_rows(read_csv(here("data/accuracy_assessment/ver3/point_probabilities_train/point_probabilities_train16_17.csv")))
```

This came from GEE: https://code.earthengine.google.com/c0e79d05ac820e225c5a816dccfd5cd9
Then get f-0.5 score across thresholds
```{r}
# run the GEE code above for thresholds between 0.1 and 0.9 by 0.10 (coarse search)
folder_path <- here("data/gee_exports/predicted_rasters/processed/confusion_matrices_across_thresholds/PPM_confusion_allPlots")

ppm_cm <- list.files(
  path = folder_path,
  pattern = "\\.csv$",
  full.names = TRUE
) |>
  map(read_csv, col_types = cols(dataset = col_character())) |>
  list_rbind()

# Ok, max f is somewhere between a 0.7 and 0.9 threshold
get_f_by_thresh(ppm_cm, 0.5)

# re-run the GEE code for thresholds between 0.7 and 0.9 by 0.01 (fine search)
folder_path <- here("data/gee_exports/predicted_rasters/processed/confusion_matrices_across_thresholds/PPM_confusion_allPlots_finesearch")

ppm_cm_fine <- list.files(
  path = folder_path,
  pattern = "\\.csv$",
  full.names = TRUE
) |>
  map(read_csv, col_types = cols(dataset = col_character())) |>
  list_rbind()

# Ok, max f is 0.77, when threshold is 0.87
get_f_by_thresh(ppm_cm_fine, 0.5)
```

# Optimal threshold for Dynamic World
Load in dynamic world's training probabilities = p(water) + p(flooded_vegetation)  
This came from GEE: https://code.earthengine.google.com/79767a9adac29835ea8f43b48f20706e
Then get f-0.5 score across thresholds
```{r}
# run the GEE code above for thresholds between 0.1 and 0.9 by 0.10 (coarse search)
dw_cm <- read_csv(here("data/gee_exports/gee_rasters/dynamicWorld/DW_confusion_allPlots.csv"))

# Ok, max f is somewhere between a 0.5 and 0.7 threshold
get_f_by_thresh(dw_cm, 0.5)

# re-run the GEE code for thresholds between 0.5 and 0.7 by 0.01 (fine search)
dw_cm_fine <- read_csv(here("data/gee_exports/gee_rasters/dynamicWorld/DW_confusion_allPlots_finesearch.csv"))

# Ok, max f is 0.749, when threshold is 0.56-0.60
get_f_by_thresh(dw_cm_fine, 0.5)
```

We've gotta pick between 0.56 and 0.60, for the dw threshold. Let's go with 0.56, because it has a better balance of recall-precision

# Explore: are ecoregion-specific threshold necessary?

```{r}
plot_ecoreg_xwalk <- st_join(plots, ecoregions, largest=T) |>
  select(plotID=Plot, L3mod) |>
  st_drop_geometry() |>
  distinct()
```

See if different decision thresholds are suggested for different ecoregions: our model
-> ecoregion-specific optimal thresholds are negligibly different from 0.87 (yay)
```{r}
ppm_cm |>
  bind_rows(ppm_cm_fine) |>
  inner_join(plot_ecoreg_xwalk) |>
  group_by(L3mod) |>
  group_modify(~ get_f_by_thresh(.x, fbeta = 0.5)) |>
  group_by(L3mod) |>
  mutate(best = f == max(f)) |>
  ggplot(aes(x=threshold, y=f, color=L3mod)) + 
  geom_line() + 
  geom_point() +
  geom_vline(xintercept = 0.87, linetype="dashed") + 
  geom_point(data = ~ filter(.x, best), size = 4, aes(fill = L3mod)) +
  labs(y="F-0.5 score", title="dynamic world")
# opt F0.5 dt's are really similar across ecoregions
```

See if different decision thresholds are suggested for different ecoregions: dynamic World
-> ecoregion-specific optimal thresholds are negligibly different from 0.56 (yay)
```{r}
dw_cm |>
  bind_rows(dw_cm_fine) |>
  inner_join(plot_ecoreg_xwalk) |>
  group_by(L3mod) |>
  group_modify(~ get_f_by_thresh(.x, fbeta = 0.5)) |>
  group_by(L3mod) |>
  mutate(best = f == max(f)) |>
  ggplot(aes(x=threshold, y=f, color=L3mod)) + 
  geom_line() + 
  geom_point() +
  geom_vline(xintercept = 0.56, linetype="dashed") + 
  geom_point(data = ~ filter(.x, best), size = 4, aes(fill = L3mod)) +
  labs(y="F-0.5 score", title="dynamic world")
```