---
title: "Process GEE raster evaluation exports"
author: "Maggie Church"
created: "2024-06-26"
updated: "2025-06-26"
---

```{r}
library(terra)
library(here)
library(dplyr)
library(sf)
library(future.apply)
library(parallel)
```

This file reads in the gee exports, combines, and saves as a compressed tif
```{r}
# read in ppr boundary
ppr <- st_read(here("data", "boundaries", "PPJV"))

# read in plot boundaries
plots <- st_read(here("data", "allPlots")) |>
  st_transform(4326)
```


# Process our classified rasters
combine and trim raw classified tifs (our model), export as 1 tif
this uses dt=0.87, and applies 10m morphological closing (conducted and exported from GEE)
```{r}
for(period in c("Aug16", "May22", "May24")){
  mosaic_raster_to_ppr(
    epsg = 32614,
    input_folder = here(paste0("data/gee_exports/predicted_rasters/raw/smooth10m/smooth10_", period)),
    output_path = here(paste0("data/gee_exports/predicted_rasters/processed/smooth10m/", period, ".tif"))
  )
}
```

# Process dynamic world rasters
this uses dt=0.30 (exported from GEE)
```{r}
for(period in c("Aug2016", "May2022", "May2024")){
  mosaic_raster_to_ppr(
    epsg = 4326,
    input_folder = here(paste0("data/gee_exports/gee_rasters/dynamicWorld/raw/", period)),
    output_path = here(paste0("data/gee_exports/gee_rasters/dynamicWorld/processed/", period, ".tif"))
  )
}
```

# DSWE - aggressive
```{r}
for(period in c("Aug2016", "May2022", "May2024")){
  mosaic_raster_to_ppr(
    epsg = 4326,
    input_folder = here(paste0("data/gee_exports/gee_rasters/dswe_aggressive/raw/", period)),
    output_path = here(paste0("data/gee_exports/gee_rasters/dswe_aggressive/processed/", period, ".tif"))
  )
}
```

# DSWE - aggressive
```{r}
for(period in c("Aug2016", "May2022", "May2024")){
  mosaic_raster_to_ppr(
    epsg = 4326,
    input_folder = here(paste0("data/gee_exports/gee_rasters/dswe_conservative/raw/", period)),
    output_path = here(paste0("data/gee_exports/gee_rasters/dswe_conservative/processed/", period, ".tif"))
  )
}
```

# JRC
```{r}
for(period in c("Aug2016", "May2022", "May2024")){
  mosaic_raster_to_ppr(
    epsg = 4326,
    input_folder = here(paste0("data/gee_exports/gee_rasters/jrc/raw/", period)),
    output_path = here(paste0("data/gee_exports/gee_rasters/jrc/processed/", period, ".tif"))
  )
}
```

# NDVI
This next chunk creates a function to split tiled NDVI exports into plot-sized tiles. 
If a plot covers multiple tiles, crops from each tile are merged to cover full plot extent. 
I don't simply mosaic these bc they're too big (perhaps bc it's a numeric var, instead of binary?)
```{r}
# set input and output paths
input_path <- here(paste0("data/gee_exports/gee_rasters/ndvi/raw/", period))
output_path <- here(paste0("data/gee_exports/gee_rasters/ndvi/processed"))

# p is a plot id
# rasters is a list of raw rasters

# Function to crop raster tiles by plots
crop_raster <- function(p, rasters, output_path) {
  
  # Get the extent for the current plot by filtering the plot_id
  plot_extent <- plots[plots$Plot == p, ]
  
  # Initialize a list to store cropped rasters
  cropped_rasters <- list()

  # Initialize a list to store cropped rasters
  for (i in seq_along(rasters)) {
    raster <- rasters[[i]]  # Extract the raster from the list
    
    # Crop the raster with error handling for non-overlapping extents
    cropped_raster <- tryCatch({
        crop(raster, plot_extent)
      }, error = function(e) {
        return(NULL)})
    
  # If a valid cropped raster was returned, store it in the list
  if (!is.null(cropped_raster) && !all(is.na(values(cropped_raster)))) {
    cropped_rasters <- c(cropped_rasters, list(cropped_raster))
    }
  }
  
  # If there are any cropped rasters, merge them or return the single raster
  if (length(cropped_rasters) > 0) {
    if (length(cropped_rasters) == 1) {
      # If there's only one cropped raster, no need to merge, just return it
      merged_raster <- cropped_rasters[[1]]
    } else {
      # Merge all cropped rasters if more than one
      merged_raster <- do.call(terra::merge, cropped_rasters)
    }
    
    # Define the output file name based on the plot ID
    output_file <- paste0(output_path, "/", period, "/raster_", p, ".tif")
    
    # Save the merged raster
    writeRaster(merged_raster, output_file, overwrite = TRUE)
    
    return(output_file)  # Return the file path of the cropped raster
  } else {
    return(NULL)  # Return NULL if no rasters were cropped
  }
}

################################################################################
## Now we run the function! I use parallel processing to speed things up... this part only takes a few minutes.

# Get unique plot IDs
plot_ids <- unique(plots$Plot)

# set up parallel processing
w <- detectCores()
plan(multisession, workers = w-2)

# Load all GEE raster exports 
files <- list.files(input_path, full.names = TRUE)
rasters <- lapply(files, rast)

# Apply the function to each plot
result_files <- lapply(plot_ids, function(p) {
  crop_raster(p, rasters, output_path=output_path)
})

# Reset the plan to sequential (stop parallelization)
plan(sequential)
```
