---
title: "Get pixel-level accuracy metrics"
author: "Maggie Church"
created: "2025-06-26"
updated: "2026-03-10"
---

**Description**: For each pixel within the historic wetland footprint across test plot-surveys,
extract observed wetland status and predictions from five remote sensing products (our PPM, 
Dynamic World, DSWE aggressive, DSWE conservative, JRC), along with covariates used in 
accuracy assessment (distances to pond edge and centroid, ndvi, phdi).

**Inputs**:
-  PPJV administrative boundary shapefile 
-  Plot boundary shapefile (created in script #1)
-  HAPET pond boundaries for all ponds (including ones < 0.04 ha) with split polygons merged (script #2)
-  Test plot-survey ids 
-  County-level PHDI from NOAA
-  Historic wetland footprint (exported from GEE)
-  Our model's probability raster (for AUC calculation) and post-processed classification raster, for each test year (exported from GEE)
-  Dynamic World, DSWE, NDVI, and JRC prediction rasters, for each test year (exported from GEE)
-  Crosswalk mapping raster values to class labels

**Steps**:
1. For each test plot-survey, identify pixels within the historic wetland footprint
2. Spatially join observed wet/dry status from HAPET pond polygons
3. Extract predicted class from each remote sensing product
4. Calculate each pixel's distance to pond edge and pond centroid
5. Extract NDVI value
6. Join PHDI via spatial join
7. Combine across all survey dates and write to CSV

**Outputs**:
- `data/accuracy_assessment/pixel_results.csv`: ~8M rows for the out-of-sample test set

```{r, message=F}
library(tidyverse)
library(here)
library(sf)
library(terra)

set.seed(123)
```

# Load data and helper function
```{r, message=F}
# Load PPR boundary
ppr <- st_read(here("data", "boundaries", "PPJV")) |> st_transform(32614)
plots <- st_read(here("data", "allPlots")) 
ponds <- st_read(here("data", "allPonds_all"))

# Add pond size 
ponds <- ponds |>
  mutate(area_ha = as.numeric(st_area(geometry))*0.0001) |>
  st_transform(4326) #ok now get into wgs84

# get test plots and time periods (we just need plot IDs)
test_plot_ids <- read_csv(here("data/train_test_data/test_plot_surveys.csv")) |> 
  distinct(plot_id) |> 
  pull()

# Load helper function to calculate PHDI
source(here("code/analysis/helper_functions/get_phdi.R")) 
```

# Define raster codes
```{r}
crosswalks <- readRDS(here("data/gee_exports/product_codes"))

crosswalks <- lapply(crosswalks, function(cw) {
  cw$value <- as.numeric(cw$value)
  cw
})
```

# Quick function to get utm zone epsg from epsg4326 sf plot polygon
```{r}
epsg_code_from_wgs <- function(plot_vect){
  # Put into projected crs for buffering - find the best UTM zone
  plot_sf <- st_as_sf(plot_vect)
  plot_centroid <- st_coordinates(st_centroid(plot_sf))
  lon <- plot_centroid[1]
  lat <- plot_centroid[2]
  zone <- floor((lon + 180) / 6) + 1
  epsg <- ifelse(lat >= 0, 32600 + zone, 32700 + zone)
  return(epsg)
}
```

# Plot to test_box lookup
```{r}
test_box_lookup <- tribble(
  ~Plot, ~test_box,
  944, 1, 950, 1, 996, 1, 992, 1, 993, 1, 937, 1, 930, 1, 940, 1, 915, 1, 905, 1,
  154, 2, 120, 2, 119, 2, 155, 2, 601, 2, 118, 2, 343, 2,
  997, 2, 998, 2, 999, 2, 1000, 2, 1001, 2, 1002, 2, 1003, 2, 1004, 2,
  768, 3, 620, 3, 204, 3, 280, 3, 281, 3, 202, 3, 201, 3, 198, 3, 282, 3, 283, 3,
  313, 4, 312, 4, 272, 4, 271, 4, 269, 4, 266, 4, 265, 4, 262, 4, 263, 4, 264, 4, 268, 4, 267, 4,
  502, 5, 69, 5, 44, 5, 178, 5, 45, 5, 50, 5, 521, 5, 340, 5, 177, 5, 416, 5, 417, 5, 621, 5, 414, 5, 412, 5,
  737, 6, 249, 6, 746, 6, 253, 6, 250, 6, 767, 6, 803, 6, 804, 6, 809, 6, 811, 6, 813, 6, 820, 6, 821, 6, 827, 6, 828, 6, 832, 6, 831, 6, 836, 6, 835, 6, 824, 6
)
```

# Helper: extract a binary prediction raster to points, rename result column
```{r}
extract_prediction <- function(rast, 
                               source, 
                               fc) {
  cw <- crosswalks[[source]]
  levels(rast) <- data.frame(ID = cw$value, category = cw$class)
  terra::extract(rast, fc, xy = FALSE) |>
    rename(!!source := category) |>
    select(-ID)
}
```

# Function to get pixel-based true + predicted values
For each (10-m) pixel within a test plot-survey and (1) within the historic wetland footprint, get: 
 (2) observed HAPET value + attributes
 (3) estimated values (our model, dw, dswe, gsw)
 (4) pixel distance to pond centroid and pond edge
 (5) NDVI 
 (6) PHDI

A note on the crs decisions:
  In this code, I avoid re-projecting rasters. I simply extract their values to the plot_pixels point collection 
      The wetlandFootprint and our PPM predictions are in utm14. Maybe that wasn't ideal, but it's not worth redoing.
      The other products' predictions and NDVI are in wgs84
  However, I do reproject observed ponds when calculating distances, using the most appropriate utm zone for that plot
```{r}
run_plot_extraction <- function(mmyy, plot_ids=test_plot_ids){
  
  # Define the mapping for dataset_label
  dataset_lookup <- c(
    "May24" = "2024",
    "May22" = "2022",
    "Aug16" = "brood16"
  )

  # Get labels for this mmyy
  label <- paste0(substr(mmyy, 1, 3), "20", substr(mmyy, 4, 5))  # e.g. May22 -> May2022
  dataset_label <- dataset_lookup[[mmyy]]            
  phdi_yr <- paste0("phdi_", tolower(mmyy))                      # e.g. May22 -> phdi_may22
  
  # Get historic wetland footprint
  wetlandFootprint <- rast(here("data/gee_exports/gee_rasters/possibilityLayer/possibilityLayer.tif"))
  
  # Read in extra files
  our_probability_rast <- rast(here(paste0("data/gee_exports/predicted_rasters/processed/probabilities/", label, ".tif")))
  postproc10_rast      <- rast(here(paste0("data/gee_exports/predicted_rasters/processed/smooth10m/", label, ".tif")))
  dw_rast              <- rast(here(paste0("data/gee_exports/gee_rasters/dynamicWorld/processed/", label, ".tif")))
  dswe_agg_rast        <- rast(here(paste0("data/gee_exports/gee_rasters/dswe_aggressive/processed/", label, ".tif")))
  dswe_cons_rast       <- rast(here(paste0("data/gee_exports/gee_rasters/dswe_conservative/processed/", label, ".tif")))
  ndvi_files           <- list.files(here(paste0("data/gee_exports/gee_rasters/ndvi/processed/", label)), pattern = "\\.tif$", full.names = TRUE)
  jrc_path             <- here(paste0("data/gee_exports/gee_rasters/jrc/processed/", label, ".tif"))
  if (file.exists(jrc_path)) jrc_rast <- rast(jrc_path)
  
  # limit to ponds within the year
  ponds_yr <- ponds |> filter(dataset == dataset_label)

  # limit to test plots 
  ponds_yr <- ponds_yr |> filter(Plot %in% plot_ids)
  plots_yr <- plots |> filter(Plot %in% plot_ids)
  
  ###################################################
  # initialize empty list
  results_list <- list()
  
  # Loop through each plot
  for(i in 1:nrow(plots_yr)) {
      
      # Load plot 
      plot_sf <- plots_yr[i,]
      plot <- vect(plot_sf)
      plot_utm <- plot |> project("EPSG:32614")   
      
      plot_id <- plot$Plot
      
      epsg <- epsg_code_from_wgs(plot_sf)
      
     ############################    
      # Get observed ponds
      # pond and plot layers align better when using WGS84 than UTM14
      true_wet <- ponds_yr |>
        filter(Plot == plot_id) |>
        st_intersection(plot_sf) |> 
        mutate(ID = row_number()) |> # need this later, to match to prob_raster extraction
        select(-Plot.1) |>
        st_make_valid() |> 
        filter(!sf::st_is_empty(geometry))

      # Skip if no observed ponds observed in this plot-survey
      if (nrow(true_wet) == 0 || all(st_is_empty(true_wet))) {
        message(paste("Skipping plot", plot_id, "- pond geometry is empty"))
        next
      }
      
       # If there were observations for this plot-survey, proceed
       if (nrow(true_wet) > 0) {
         
         print(paste("processing plot", plot_id, "- survey", mmyy))
         
          # Buffer 10m to avoid weird slivers (use utm14 for this, then put back into wgs84)
          plot_extent_utm <- terra::buffer(plot_utm, width = -10) 
          plot_extent <- plot_extent_utm |> project("EPSG:4326") 
          
         ###################
          # 1. Get a feature collection of points to compare predictions to observations: 10-m pixel centroids from historic wetland layer
          plot_pixels_utm <- crop(wetlandFootprint, plot_extent_utm) |>  
            clamp(lower = 0, upper = 1, values = FALSE) |>                 
            as.points() |>
            st_as_sf() |>
            filter(possiblyWet == 1) |>
            st_intersection(st_as_sf(plot_extent_utm)) |>  # clip to actual plot shape, not just bbox 
            select(-c(WMD, Shp_Lng, Shap_Ar)) 
          
           plot_pixels <- st_transform(plot_pixels_utm, 4326)
          
         ###################
          # 2 Get observed values to plot_pixels
          # Get the dry areas from the complement of the wet areas
          true_dry <- st_difference(st_geometry(st_as_sf(plot_extent)), st_union(true_wet)) |> 
            st_sf() |>
            rename(geometry=1) |>
            filter(!st_geometry_type(geometry) %in% c("MULTILINESTRING", "LINESTRING")) |>
            st_make_valid() 
          
           # Add dry attributes 
           dry_values <- true_dry |> 
             mutate(true_label = "dry",
                    dataset = dataset_label) 
           
          # Add wet attributes (location, size, veg)
          wet_values <- true_wet |> 
            st_intersection(st_as_sf(plot_extent)) |> # crop true_wet to the in-buffered plot too
            select(ID, dataset, area_ha, pond_id=rownum) |> 
            mutate(true_label = "wet")
                     
          # combine 
          observed <- bind_rows(dry_values, wet_values)
          
          # Add true label via spatial join
          observed_values <- st_join(plot_pixels, observed) |>
            distinct(geometry, .keep_all = TRUE)
          
         ###################
         # 3. Get predicted values for this plot
         #    (Don't reproject prediction rasters, it's best not to reproject  
         #    rasters where we can help it, plus it takes forever)
         probabilities     <- terra::extract(our_probability_rast, plot_pixels_utm, xy = FALSE) |> select(-ID)
         postproc10_values <- extract_prediction(postproc10_rast, "postproc10",  plot_pixels_utm)
         dw_values         <- extract_prediction(dw_rast,         "dw",          plot_pixels)
         dswe_agg_values   <- extract_prediction(dswe_agg_rast,   "dswe_agg",    plot_pixels)
         dswe_cons_values  <- extract_prediction(dswe_cons_rast,  "dswe_cons",   plot_pixels)

         # combine estimates from different products
         plot_results <- bind_cols(
           observed_values,
           probabilities,
           postproc10_values, 
           dw_values,
           dswe_agg_values,
           dswe_cons_values)
         
        # add jrc if it exists (aug16)
        if (file.exists(jrc_path)) {
          jrc_values <- extract_prediction(jrc_rast, "jrc", plot_pixels)
          plot_results <- bind_cols(plot_results, jrc_values)
        }
         
         ###################
         # 4a. Calculate distance to pond centroids
         true_wet_utm <- st_transform(true_wet, epsg)
         
         pond_centroids <- st_centroid(true_wet_utm) |> 
            select(ID, centroid_geom=geometry) |> 
            as.data.frame(xy=T)
                          
         # Add distance from point to pond centroid
         plot_results <- plot_results |>
            st_transform(epsg) |>
            left_join(pond_centroids, by = "ID") |>
            mutate(dist_to_centroid_m = if_else(
                true_label == "wet",
                as.numeric(st_distance(geometry, centroid_geom, by_element = TRUE)),
                NA_real_
            )) |>
            select(-centroid_geom) |>
            st_transform(4326)
         
         ###################
         # 4b. Calculate distance to pond edges
          true_dry_utm  <- st_transform(true_dry, epsg)
          template_rast <- crop(wetlandFootprint, plot_extent_utm)
          
          wet_rast <- terra::rasterize(vect(true_wet_utm), template_rast, field = 1)
          dry_rast <- terra::rasterize(vect(true_dry_utm), template_rast, field = 1)
          
          dry_dist_wgs <- project(terra::distance(wet_rast), "EPSG:4326")
          wet_dist_wgs <- project(terra::distance(dry_rast), "EPSG:4326")
          
          plot_results <- plot_results |>
            st_transform(epsg) |>
            mutate(dist_to_edge_m = if_else(
              true_label == "wet",
              terra::extract(wet_dist_wgs, plot_pixels)[, 2],
              terra::extract(dry_dist_wgs, plot_pixels)[, 2]
            )) |>
            st_transform(4326)
          
         ###################
         # 5. NDVI
         ndvi_file <- ndvi_files[str_detect(ndvi_files, paste0("_", plot_id, "\\.tif$"))]
          
         if (length(ndvi_file) == 0) {
           message(paste("Skipping plot", plot_id, "- NDVI file not found."))
           next
         }
          
         plot_results <- bind_cols(
           plot_results,
           terra::extract(rast(ndvi_file), plot_pixels, xy = FALSE) |> select(-ID)
           )
         
         ###################
         # Store result
         results_list[[i]] <- plot_results |>
           mutate(x = st_coordinates(geometry)[, 1],
                  y = st_coordinates(geometry)[, 2]) 
    }
  } # end of plot-level stuff
  
  #############################################
   # Combine all results into one data frame
  point_probabilities <- do.call(rbind, results_list) |>
    mutate(STATE = if_else(dataset %in% c("brood16", "pair16", "brood17", "pair17"), "ND", STATE)) |>
    filter(!is.na(STATE)) |>
    left_join(test_box_lookup)

  # 6. Add PHDI via spatial join
  processed_phdi_list <- lapply(processed_phdi_list, function(layer) {
    st_transform(layer, 4326) |> select(PHDI)
  })

  point_probabilities <- point_probabilities |>
    st_join(processed_phdi_list[[phdi_yr]], left = FALSE) |>
    st_drop_geometry()

  return(point_probabilities)
}
```

# Run function
This creates a dataset for every non-clouded pixel with the historic wetland footprint, with our model's estimated probabilities and classifications from other models (water, nonwater, or NA for clouded/noData)
```{r}
# Get OOST set
point_probabilities_May22 <- run_plot_extraction(mmyy="May22")
point_probabilities_May24 <- run_plot_extraction(mmyy="May24")
point_probabilities_Aug16 <- run_plot_extraction(mmyy="Aug16")

point_probabilities_oost <-
  bind_rows(
    point_probabilities_May22,
    point_probabilities_May24,
    point_probabilities_Aug16
  )

# this has 8 mil oost points in the end! 
point_probabilities_oost |> 
  write_csv(here(paste0("data/accuracy_assessment/pixel_results.csv")))
```
