---
title: "get plot results"
author: "Maggie Church"
created: "2025-10-22"
updated: "2026-03-25"
---

**Description**:  Evaluates the plot-level accuracy of surface water classification products 
(our post-processed model, Dynamic World, DSWE aggressive, DSWE conservative, and JRC) 
against HAPET ground-truth pond observations across PPJV survey plots. For each 
out-of-sample plot-survey, computes a) predicted vs. observed inundated area by class (water, dry, cloud, nonwetland) 
and b) IoU-based pond detection metrics stratified by pond size class (<0.5 ha, 0.5–4 ha, >4 ha).

**Inputs**:
-  PPJV administrative boundary shapefile 
-  Plot boundary shapefile (created in script #1)
-  HAPET pond boundaries for all ponds (including ones < 0.04 ha) with split polygons merged (script #2)
-  Test plot-survey ids (we can stratify these later)
-  County-level PHDI from NOAA
-  Prediction rasters for each test year (exported from GEE): our post-processed classification raster, Dynamic World, DSWE, NDVI, and JRC
-  Crosswalk mapping raster values to class labels

**Steps**:
1. Read in plot boundaries, HAPET pond observations, and prediction rasters for all products
2. For each out-of-sample plot-survey:
   a. Clip plot to a 1000 ha square in the appropriate UTM zone
   b. Spatially join PHDI drought index to observed ponds
   c. For each product, crop prediction raster to plot extent and compute area by class (water, dry, cloud)
   d. Compute cloud-adjusted true wet area
   e. Calculate IoU for each observed pond vs. predicted water, stratified by pond size class
3. Combine results across three test years (Aug 2016, May 2022, May 2024)
4. Exclude plots with unreliable HAPET digitizations (plots 204, 620)

**Outputs**:
- `polygon_assessment_oost_plots_post_processed.csv` — plot-level area metrics by product and year
- `iou_assessment_oost_plots_post_processed.csv` — pond-level IoU metrics by size class, product, and year

```{r}
library(tidyverse)
library(here)
library(terra)
library(sf)

options(sf_use_s2 = TRUE)
```

# Read in datasets
```{r}
ppr <- st_read(here("data", "boundaries", "PPJV")) |> st_transform(32614)
plots <- st_read(here("data", "allPlots")) 
ponds <- st_read(here("data", "allPonds_all")) |> st_transform(4326)

# convert plots to a SpatVector
plots_terra <- vect(plots)

# set test plots and time periods
test_plot_ids <- read_csv("data/train_test_data/test_plot_surveys.csv")

# get PHDI
source(here("code/analysis/helper_functions/get_phdi.R"))

phdi <- do.call(rbind, lapply(names(phdi_data_list), function(n) {
  cbind(source = n, processed_phdi_list[[n]])
})) |>
  mutate(PHDI = as.numeric(PHDI)) |>
  filter(!is.na(PHDI))
```

# Read in estimation rasters
```{r}
# load post-proc images
postproc_files10 <- list.files(path = here("data/gee_exports/predicted_rasters/processed/smooth10m/"), pattern = "\\.tif$", full.names = TRUE)
postproc_rasters10 <- lapply(postproc_files10, terra::rast)
names(postproc_rasters10) <- tools::file_path_sans_ext(basename(postproc_files10))

# load jrc images
jrc_files <- list.files(path = here("data/gee_exports/gee_rasters/jrc/processed"), pattern = "\\.tif$", full.names = TRUE)
jrc_rasters <- lapply(jrc_files, terra::rast)
names(jrc_rasters) <- tools::file_path_sans_ext(basename(jrc_files))

# load dw images
dw_files <- list.files(path = here("data/gee_exports/gee_rasters/dynamicWorld/processed"), pattern = "\\.tif$", full.names = TRUE)
dw_rasters <- lapply(dw_files, terra::rast)
names(dw_rasters) <- tools::file_path_sans_ext(basename(dw_files))

# load dswe-aggressive images
dswe_agg_files <- list.files(path = here("data/gee_exports/gee_rasters/dswe_aggressive/processed"), pattern = "\\.tif$", full.names = TRUE)
dswe_agg_rasters <- lapply(dswe_agg_files, terra::rast)
names(dswe_agg_rasters) <- tools::file_path_sans_ext(basename(dswe_agg_files))

# load dswe-conservative images
dswe_cons_files <- list.files(path = here("data/gee_exports/gee_rasters/dswe_conservative/processed"), pattern = "\\.tif$", full.names = TRUE)
dswe_cons_rasters <- lapply(dswe_cons_files, terra::rast)
names(dswe_cons_rasters) <- tools::file_path_sans_ext(basename(dswe_cons_files))
```

# Define raster codes
```{r}
crosswalks <- readRDS(here("data/gee_exports/product_codes"))

crosswalks <- lapply(crosswalks, function(cw) {
  cw$value <- as.numeric(cw$value)
  cw
})
```

# Quick function to get utm zone epsg from epsg4326 sf plot polygon
```{r}
epsg_code_from_wgs <- function(plot_vect){
  # Put into projected crs for buffering - find the best UTM zone
  plot_sf <- st_as_sf(plot_vect)
  plot_centroid <- st_coordinates(st_centroid(plot_sf))
  lon <- plot_centroid[1]
  lat <- plot_centroid[2]
  zone <- floor((lon + 180) / 6) + 1
  epsg <- ifelse(lat >= 0, 32600 + zone, 32700 + zone)
  return(epsg)
}
```

# Make plot square function
```{r}
# this'll ensure the plot is a 1000 ha square
make_square <- function(obj_wgs, area, epsg) {
  
  # Put into projected crs for buffering - find the best UTM zone
  plot_sf_utm <- st_transform(obj_wgs, epsg)
  
  # make square
  bbox <- st_bbox(plot_sf_utm)
  width  <- bbox["xmax"] - bbox["xmin"]
  height <- bbox["ymax"] - bbox["ymin"]
  
  if (width > height) {
    diff = width - height
    bbox["xmax"] <- bbox["xmax"] - diff
  } else if (height > width) {
    diff = height - width
    bbox["ymax"] <- bbox["ymax"] - diff
  }
  
  square_poly <- st_as_sfc(bbox)
  plot_square <- st_intersection(plot_sf_utm, square_poly)
  
  # the resulting square isn't axis-aligned, so I can't assume area=l*w
  # I have to iteratively find the buffer that makes the square 1000 ha
  f <- function(d) {abs(as.numeric(st_area(st_buffer(plot_square, dist=-d))) - area)}
  opt_d <- optimize(f, interval=seq(0, 50, 0.5))$minimum
  plot_sf_utm_buff <- st_buffer(plot_square, dist=-opt_d)
  
  # convert back to wgs84
  plot_sf_buff <- st_transform(plot_sf_utm_buff, 4326)
  return(plot_sf_buff)
}
```

# Fn to calculate IOU metrics
```{r}
calculate_iou <- function(true_polys, pred_polys){
   inter <- st_intersection(true_polys, pred_polys)
   union <- st_union(bind_rows(true_polys, pred_polys))
   
   inter_area <- sum(st_area(inter))  
   union_area <- sum(st_area(union))  
   iou <- as.numeric(inter_area/union_area)
   
   return(iou)
}
```

# area counts by class
Function to, for a given product:
 1. crop a ppr-wide raster into a plot-size chunk
 2. convert to UTM and crop again so that the raster is exactly a 100 ha square 
 3. get area for each class (water, dry, cloud)
 4. get corresponding true HAPET wet area for noncloudy areas

Inputs: 
  pred_raster: a ppr-wide prediction raster, with categories defined in the crosswalk
  source: name of product. We'll use this for the crosswalk so it needs to be a specific name
  plot: original plot boundary (vect, 4326)
  plot_square: plot boundary cropped to be 100 ha and square-shaped (sf, 4326)
  true_wet: true wet HAPET polygons
  epsg: UTM zone for cropping and calculating area

Outputs: df with 
 - Surface water extent (observed - clouds)
 - Surface water extent (estimated)
 - By pond size class (<0.5ha, 0.5-2ha, >2ha): mean IoU, # ponds "detected" at different IoU thresholds (0, 0.25, 0.5)
 - # observed ponds
```{r}
count_area <- function(pred_raster, source, plot, plot_square, true_wet, epsg){
  
  # Skip if input is NULL or not a raster
  if (is.null(pred_raster) || !inherits(pred_raster, "SpatRaster")) {
    # Return empty data frames instead of NULL
    plot_metrics <- tibble()
    iou_summary <- tibble(
      size_class = c("small", "medium", "large"),
      mean_iou   = NA,
      n_gt_0.5   = NA,
      n_gt_0.25  = NA,
      n_gt_0     = NA,
      n          = NA
    )
    return(list(plot_metrics = plot_metrics, iou_summary = iou_summary))
  }  
  # what epsg is the prediction raster in?
  crs_pred_raster <- crs(pred_raster, describe = TRUE)$code
  
  # Crop the ppr raster to the plot first, in the raster's epsg
  if (crs_pred_raster != 4326) {
    plot_extent <- plot |> terra::project(paste0("EPSG:", crs_pred_raster)) |> terra::ext()
  } else {
    plot_extent <- terra::ext(plot)
  }
  
  # crop
  cropped_pred_rast <- terra::crop(pred_raster, plot_extent)
  names(cropped_pred_rast) <- "value"
  cropped_pred_rast <- as.factor(cropped_pred_rast)
  
  # get needed layers into UTM
  plot_square_utm <- sf::st_transform(plot_square, epsg)
  true_wet_utm <- st_transform(true_wet, epsg)
  
  # project to utm and make it match the square 100 ha plot exactly
  cropped_pred_sf_utm <- 
    cropped_pred_rast |> 
    terra::project(paste0("EPSG:", epsg)) |>
    terra::as.polygons() |> 
    sf::st_as_sf() |> 
    sf::st_intersection(plot_square_utm)
  
  # calculate area by class
  areas <- cropped_pred_sf_utm |> 
    select(value) |> 
    mutate(
      pred_area_m2 = as.numeric(st_area(geometry)),
      value = as.numeric(as.character(value))) |>
    left_join(crosswalks[[source]], by = "value")
  
  # true water surface area of the plot-survey (based only on HAPET data)
  true_wet_area_m2 <- true_wet_utm |> 
    st_intersection(plot_square_utm) |>                  # 100 ha square plot
    mutate(actual_area = as.numeric(sf::st_area(geometry))) |> 
    summarize(true_wet_area_m2 = sum(actual_area)) |>    # return total water surface area
    pull(true_wet_area_m2)
    
  # if there are clouds in this plot, get cloud-adjusted true water surface area
  if("clouds" %in% areas$class){
    
    # get cloud mask
    clouds <- areas |> filter(class=="clouds")
    
    # if the entire plot is clouded, set cloud-adjusted true water surface area = 0
    # allows a rounding area of 1 m2
    if(abs(pull(clouds, pred_area_m2) - 1e7) < 1){
      true_wet_area_m2_cloudless <- 0
      } else {
    
    # total actual water surface area of the plot-survey, where this product is cloudless
    true_wet_area_m2_cloudless <- 
      true_wet_utm |> 
      st_intersection(plot_square_utm) |>                                 # 100 ha square plot
      st_difference(clouds) |>                                            # remove clouds
      st_collection_extract("POLYGON", warn = FALSE) |>                   # remove slivers (non-polygons)
      mutate(true_cloudless_area = as.numeric(st_area(geometry))) |>      # remove slivers (too small)
      filter(true_cloudless_area > 1) |>                                  # remove slivers  (too small)
      summarize(true_wet_area_m2_cloudless = sum(true_cloudless_area)) |> # return true wet, cloudless areas
      pull(true_wet_area_m2_cloudless)
    }
  } else {
    true_wet_area_m2_cloudless <- true_wet_area_m2
  }
  
  plot_metrics <- areas |> 
    st_drop_geometry() |> 
    mutate(true_wet_area_m2_cloudless = true_wet_area_m2_cloudless,
           true_wet_area_m2 = true_wet_area_m2,
           source = source)
  
  ##########################################
  # I'm going to calculate IoU while we're here
  
    ### get cloudless true water ###
    if("clouds" %in% areas$class){
      
      clouds <- areas |> filter(class=="clouds")                          # get cloud mask
      
      true_wet_utm_cloudless <-  true_wet_utm |> 
        st_intersection(plot_square_utm) |>                               # limit to 1000 ha
        st_difference(clouds) |>                                          # remove clouds
        st_collection_extract("POLYGON", warn = FALSE) |>                 # remove slivers (non-polygons)
        mutate(true_cloudless_area = as.numeric(st_area(geometry))) |>    # remove slivers (too small)
        filter(true_cloudless_area > 1)                                    # remove slivers  (too small)
      
    } else {
        true_wet_utm_cloudless <- true_wet_utm |> 
          st_intersection(plot_square_utm)
    }

  ### limit predictions to water ### 
  pred_water_sf_utm <- cropped_pred_sf_utm |> 
    mutate(value = as.numeric(as.character(value))) |>    
    left_join(crosswalks[[source]], by = "value") |> 
    filter(class=="water") |> 
    st_cast("POLYGON") 
    
  # calculate overall plot IoU
  true_wet_small <- true_wet_utm_cloudless |> filter(as.numeric(st_area(geometry)) < 5000) # < 0.5 ha
  true_wet_med <- true_wet_utm_cloudless |> filter(as.numeric(st_area(geometry)) >= 5000 & as.numeric(st_area(geometry)) <= 40000) # 0.5-4 ha
  true_wet_lar <- true_wet_utm_cloudless |> filter(as.numeric(st_area(geometry)) > 40000) # > 4 ha
  
  # Initialize empty vector 
  calc_ious <- function(true_wet_polys){
    
    # loop through each true pond and calculate IOU
    ious <- numeric(nrow(true_wet_polys))
    
    for(i in 1:nrow(true_wet_polys)){
      true_poly <- true_wet_polys[i, ]
      pred_poly <- pred_water_sf_utm |> filter(lengths(st_intersects(pred_water_sf_utm, true_poly)) > 0)
      ious[i] <- calculate_iou(true_poly, pred_poly)
    }
    
    # combine pond ious
    list(
      mean_iou = mean(ious),
      n_gt_0.5 = sum(ious > 0.5),
      n_gt_0.25 = sum(ious > 0.25),
      n_gt_0 = sum(ious > 0), # any detection
      n = nrow(true_wet_polys)
    )
  }
  
  iou_small <- calc_ious(true_wet_small)
  iou_med <- calc_ious(true_wet_med)
  iou_lar <- calc_ious(true_wet_lar)
  
  # Combine into a single table
  iou_summary <- bind_rows(
    tibble(size_class = "small", !!!iou_small),
    tibble(size_class = "medium", !!!iou_med),
    tibble(size_class = "large", !!!iou_lar)
  )
  
  ########################################
  list(
    plot_metrics = plot_metrics,
    iou_summary  = iou_summary
  )
}
```


# Get actual + estimated results at pond and plot levels
I recognize this fn is a monster. I just kept adding to it...

This calls on each of the functions above (for a given year)
 epsg_code_from_wgs()
 make_square() 
 count_area()
```{r}
evaluate_plot_inundation <- function(year_label, 
                                    mmYYYY,
                                    phdi_yr = NULL,
                                    oos = TRUE, 
                                    area = 1e7){
  
  # limit to ponds within the year
  ponds_yr <- ponds |> filter(dataset==year_label) 
  
  # keep only test plots (optional)
  if(oos==T){
    ponds_yr <- ponds_yr |> filter(Plot %in% test_plot_ids$plot_id)
  }
  
  # initialize empty list
  plot_level_list <- list()
  iou_list <- list()
  
  for(i in 1:nrow(plots)){

      ##############################################
      # 1. Get pixel-based actual + estimated results for plot i:
      # Load plot 
      plot <- plots_terra[i,] 
      plot_id <- plot$Plot
      message("Processing plot: ", plot_id)
      
      # get UTM epsg code for this plot (we'll use this for clipping, buffering, and area calculations)
      plot_sf <- st_as_sf(plot)
      epsg <- epsg_code_from_wgs(plot_sf)
      
      # clip plot to be a 1000 ha square - we'll use this as a cookie cutter for 
      # all layers, to ensure they're totally aligned
      plot_sf_square <- make_square(plot_sf, area = 1e7, epsg=epsg)
      
      # Get actual ponds for this plot (true observations)
      true_wet <- ponds_yr |> filter(Plot == plot_id)
      
      # If an actual ponds are found within this plot, get predicted metrics
      if (nrow(true_wet) > 0) {
        
        # If there are weird slivers in true_wet, deal with it here
        if(any(!st_is_valid(true_wet))){
          true_wet <- true_wet |>
            sf::st_make_valid()
        } 
        
    ##############################################
    # 2. Add PHDI
    phdi_filtered <- phdi |> 
      filter(source == phdi_yr) |> 
      st_transform(4326)
    
    # Spatial join to get PHDI to the true_wet sf (if a pond is inter-county, 
    # it'll have 2 PHDIs. Get the weighted mean)
    pond_phdis <- st_intersection(true_wet, phdi_filtered[c("PHDI", "name")]) |>
      mutate(intersect_area = as.numeric(st_area(geometry))) |>
      group_by(rownum) |>
      summarize(
        PHDI = sum(PHDI * intersect_area, na.rm = TRUE) / sum(intersect_area, na.rm = TRUE),
        .groups = "drop"
      ) |>
      st_drop_geometry() 
    
    true_wet <- left_join(true_wet, pond_phdis, by = "rownum")
    
    # get PHDI (averaged over all ponds) for this plot
    plot_phdi <- true_wet |> 
      distinct(PHDI) |> 
      summarize(mean(PHDI)) |> 
      pull()
    
    ##############################################
    # 3. Plot level evaluation
    if (any(st_geometry_type(true_wet) == "GEOMETRYCOLLECTION")) {
      true_wet <- true_wet |>
        st_collection_extract("POLYGON", warn = FALSE)
    } 
    
    # jrc doesn't go past 2021; all others are available for our study period
    area_counts_jrc       <- count_area(jrc_rasters[[mmYYYY]], source="jrc",
                                        plot, plot_sf_square, true_wet, epsg=epsg)
    area_counts_dw        <- count_area(dw_rasters[[mmYYYY]], source="dw",
                                        plot, plot_sf_square, true_wet, epsg=epsg)
    area_counts_dswe_agg  <- count_area(dswe_agg_rasters[[mmYYYY]], source="dswe_agg",
                                        plot, plot_sf_square, true_wet, epsg=epsg)
    area_counts_dswe_cons <- count_area(dswe_cons_rasters[[mmYYYY]], source="dswe_cons",
                                        plot, plot_sf_square, true_wet, epsg=epsg)
    area_counts_postproc10 <- count_area(postproc_rasters10[[mmYYYY]], source="postproc10",
                                        plot, plot_sf_square, true_wet, epsg=epsg)
    
  # Combine plot-level area metrics
  plot_results <- dplyr::bind_rows(
    area_counts_postproc10$plot_metrics,
    area_counts_jrc$plot_metrics,
    area_counts_dw$plot_metrics,
    area_counts_dswe_agg$plot_metrics,
    area_counts_dswe_cons$plot_metrics
  ) |> 
    mutate(Plot = plot_id,
           dataset = year_label,
           phdi = plot_phdi)
  
  # Combine IoU summaries
  iou_results <- dplyr::bind_rows(
    area_counts_postproc10$iou_summary  |> mutate(source="postproc10"),
    area_counts_jrc$iou_summary       |> mutate(source="jrc"),  
    area_counts_dw$iou_summary        |> mutate(source="dw"),
    area_counts_dswe_agg$iou_summary  |> mutate(source="dswe_agg"),
    area_counts_dswe_cons$iou_summary |> mutate(source="dswe_cons")
  ) |> 
    mutate(Plot = plot_id,
           dataset = year_label)
    
    ##############################################
    # Store results in list
    plot_level_list[[i]] <- plot_results
    iou_list[[i]] <- iou_results

  } else {
    message("Skipping plot ", plot_id, " — prediction raster has no wet areas")
    next  
  }
}

# Combine plot-level results
all_plot_results <- do.call(bind_rows, plot_level_list)
all_ious_results <- do.call(bind_rows, iou_list)

# Add test box id
all_plot_results <- all_plot_results |> 
  mutate(test_box = case_when(
    Plot %in% c(944, 950, 996, 992, 993, 937, 930, 940, 915, 905) ~ 1,
    Plot %in% c(154,120,119,155,601,118,343, 997:1004) ~ 2, 
    Plot %in% c(768,620,204,280,281,202,201,198,282,283) ~ 3, 
    Plot %in% c(313,312,272,271,269,266,265,262,263,264,268,267) ~ 4,
    Plot %in% c(502,69,44,178,45,50,521,340,177,416,417,621,414,412) ~ 5,
    Plot %in% c(737,249,746,253,250,767,803,804,809,811,813,820,821,827,828,832,831,836,835,824) ~ 6,
    T ~ NA
  ))  

results <- list(
   all_plot_results = all_plot_results,
   all_ious_results  = all_ious_results
)

return(results)
}
```

Execute function for OOST sets
```{r}
oost_aug16 <- evaluate_plot_inundation(
  year_label = "brood16",
  mmYYYY = "Aug2016",
  phdi_yr = "phdi_aug16",
  oos=T
)

oost_may22 <- evaluate_plot_inundation(
  year_label = "2022",
  mmYYYY = "May2022",
  phdi_yr = "phdi_may22",
  oos=T
)

oost_may24 <- evaluate_plot_inundation(
  year_label = "2024",
  mmYYYY = "May2024",
  phdi_yr = "phdi_may24",
  oos=T
)

########
all_plot_results <- bind_rows(oost_aug16$all_plot_results, oost_may22$all_plot_results, oost_may24$all_plot_results)
all_iou_results <- bind_rows(oost_aug16$all_ious_results, oost_may22$all_ious_results, oost_may24$all_ious_results)

write_csv(all_plot_results, here(paste0("data/accuracy_assessment/plot_results.csv")))
write_csv(all_iou_results, here(paste0("data/accuracy_assessment/iou_results.csv")))
```
