---
title: "Get estimated pond polygons"
author: "Maggie Church"
created: "2025-12-30"
updated: "2026-03-25"
---

**Description**: Evaluates pond detection accuracy of five surface water classification 
products (our post-processed model, Dynamic World, DSWE aggressive, DSWE conservative, 
and JRC) against HAPET ground-truth pond observations across out-of-sample test plots. 
For each plot-survey, crops prediction rasters and observed pond polygons to a standardized 
1000 ha square plot boundary, then exports all observed, estimated, and cloud polygons 
as a single shapefile for downstream analysis.

**Inputs**:
-  Plot boundary shapefile (created in script #1)
-  HAPET pond boundaries for all ponds (including ones < 0.04 ha) with split polygons merged (script #2)
-  Test plot-survey ids (we can stratify these later)
-  Prediction rasters for each test year (exported from GEE): our post-processed classification raster, Dynamic World, DSWE, NDVI, and JRC

**Steps**:
1. Load test plot and pond data; filter to out-of-sample plot-surveys
2. For each plot, compute the appropriate UTM zone and crop to a standardized 1000 ha square boundary
3. For each product and year, crop prediction rasters to each test plot and extract water and cloud polygons (8-neighbor rule)
4. Crop observed HAPET ponds to each test plot square
5. Combine observed ponds, estimated water polygons, and cloud masks across all products and years

**Notes**:
- Plot 253 / dataset 2022 excluded — suspected incomplete HAPET delineation of river
- All plots standardized to UTM Zone 14N (EPSG:32614) for area calculations
- JRC only available for Aug 2016

```{r}
library(tidyverse)
library(sf)
library(terra)
library(here)
```

```{r setup, include=FALSE}
plots <- st_read(here("data/allPlots")) # keep in wgs84 for now

# observed ponds
ponds <- st_read(here("data", "allPonds_all")) 

# set test plots and time periods
test_plot_ids <- read_csv("data/train_test_data/test_plot_surveys.csv") |> 
  rename(Plot = plot_id)

# actually kill this plot-year. I think maybe HAPET didn't delineate the river, hard to prove
test_plot_ids <- test_plot_ids |> filter(!(Plot==253 & dataset=="2022"))

# postprocessed raster
may2022_pp1 <- rast(here("data/gee_exports/predicted_rasters/processed/smooth10m/May2022.tif"))
may2024_pp1 <- rast(here("data/gee_exports/predicted_rasters/processed/smooth10m/May2024.tif"))
aug16_pp1 <- rast(here("data/gee_exports/predicted_rasters/processed/smooth10m/Aug2016.tif"))

jrc_Aug16 <- rast(here("data/gee_exports/gee_rasters/jrc/processed/Aug2016.tif"))

dw_Aug16 <- rast(here("data/gee_exports/gee_rasters/dynamicWorld/processed/Aug2016.tif"))
dw_May22 <- rast(here("data/gee_exports/gee_rasters/dynamicWorld/processed/May2022.tif"))
dw_May24 <- rast(here("data/gee_exports/gee_rasters/dynamicWorld/processed/May2024.tif"))

dswe_agg_Aug16 <- rast(here("data/gee_exports/gee_rasters/dswe_aggressive/processed/Aug2016.tif"))
dswe_agg_May22 <- rast(here("data/gee_exports/gee_rasters/dswe_aggressive/processed/May2022.tif"))
dswe_agg_May24 <- rast(here("data/gee_exports/gee_rasters/dswe_aggressive/processed/May2024.tif"))

dswe_cons_Aug16 <- rast(here("data/gee_exports/gee_rasters/dswe_conservative/processed/Aug2016.tif"))
dswe_cons_May22 <- rast(here("data/gee_exports/gee_rasters/dswe_conservative/processed/May2022.tif"))
dswe_cons_May24 <- rast(here("data/gee_exports/gee_rasters/dswe_conservative/processed/May2024.tif"))
```

## Get obs and predicted ponds for test set
Limit to ponds and plots within OOST test set. Also calculate size classes while we're here.
Eh I'm just gonna skip 2016. Makes evaluation of rolling means weird (bc we don't estimate 2015), and makes results weird to interpret
```{r}
# filter ponds to test plots
test_ponds <- inner_join(ponds, test_plot_ids) |> st_transform(32614)

# filter plots to test plots
test_plots_sf <- inner_join(plots, test_plot_ids)
```

fn to make plot square 
```{r}
# Quick function to get the appropriate utm zone for each plot (rn they're in wgs84),
# based on plot centroid. Add the epsg code to a new column
get_utm_epsg <- function(sf_obj) {
  plot_centroid <- st_coordinates(st_centroid(sf_obj))
  lon <- plot_centroid[1]
  lat <- plot_centroid[2]
  zone <- floor((lon + 180) / 6) + 1
  epsg <- ifelse(lat >= 0, 32600 + zone, 32700 + zone)
  return(epsg)
}

# apply fn
test_plots_sf <- test_plots_sf |>
  mutate(
    utm_epsg = sapply(seq_len(n()), function(i) get_utm_epsg(test_plots_sf[i, ]))
  )
```

fn to make each plot a 1000 ha square, to deal with some weird edge effects 
```{r}
make_square <- function(obj_wgs, area, epsg) {
  
  # Put into projected crs for buffering - find the best UTM zone
  plot_sf_utm <- st_transform(obj_wgs, epsg)
  
  # Get bbox height and width 
  bbox <- st_bbox(plot_sf_utm)
  width  <- bbox["xmax"] - bbox["xmin"]
  height <- bbox["ymax"] - bbox["ymin"]
  
  # Make the bbox square, use the length of whichever side is shorter
  if (width > height) {
    diff = width - height
    bbox["xmax"] <- bbox["xmax"] - diff
  } else if (height > width) {
    diff = height - width
    bbox["ymax"] <- bbox["ymax"] - diff
  }
  
  # Crop the plot sf by the square bbox
  square_poly <- st_as_sfc(bbox, crs = st_crs(plot_sf_utm))
  plot_square <- st_intersection(plot_sf_utm, square_poly)
  
  # Now make this square plot 1000 ha
  # The plots aren't axis-aligned (pointing north), so I can't assume area=l*w
  # I have to iteratively find the buffer that makes the square 1000 ha
  f <- function(d) {abs(as.numeric(st_area(st_buffer(plot_square, dist=-d))) - area)}
  opt_d <- optimize(f, interval=seq(0, 50, 0.5))$minimum
  plot_sf_utm_buff <- st_buffer(plot_square, dist=-opt_d)
  
  # Convert to UTM14, so that I don't have to re-project any estimated rasters 
  # to UTM15 (which may change its landscapemetrics). Bc I'm most interested in
  # pond counts, this should be fine. 
  plot_sf_wgs_buff <- st_transform(plot_sf_utm_buff, 32614)
  
  return(plot_sf_wgs_buff)
}

# apply fn
test_plots_sf_square <- do.call(
  rbind, 
  lapply(seq_len(nrow(test_plots_sf)), function(i) {
    make_square( 
      obj_wgs = test_plots_sf[i, ], 
      area = 1000 * 1e4, 
      epsg = test_plots_sf$utm_epsg[i] 
      ) 
    }) 
  )
```

fn to calculate pond metrics for a plot-survey (eg #ponds) and also return obs and est polygons
```{r}
# Function to crop raster tiles by plots
get_pond_metrics <- function(plot_idx, raster, dataset_sel, cloudval, waterval) {
  
  plot_square <- test_plots_sf_square[plot_idx, ]
 
  ################################################
  # Crop estimated raster to current square plot

  # if raster is in wgs84, then reproject the plot ext before cropping
  # our model and the plot ext are both in utm14 otherwise
  if(crs(raster, describe = TRUE)$code == 4326){
     plot_square_rastcut <- plot_square |> st_transform(4326)
  } else {
    plot_square_rastcut <- plot_square
  }
  
  ponds_est <- crop(raster, vect(plot_square_rastcut))
  
  # binary water/nonwater raster
  ponds_est_bin <- ponds_est
  ponds_est_bin[ponds_est_bin != waterval] <- NA
  
  # cloud raster
  clouds <- ponds_est
  clouds[clouds != cloudval] <- NA
  
  # I could convert to vector, or use landscapemetrics. I don't think it matters, 
  # and I'd rather be able to sanity check with vectors
  ponds_est_sf <- terra::as.polygons(
    ponds_est_bin,
    dissolve = TRUE,   
    directions = 8      # 8-neighbor rule
  ) |> 
    st_as_sf() |> 
    st_cast("POLYGON") |>
    st_intersection(plot_square_rastcut) |> #crop to square plot
    mutate(pond_size = as.numeric(st_area(geometry)))
  
  ################################################
  # add clouds
  clouds_sf <- clouds |>
    terra::as.polygons(dissolve = TRUE) |>
    sf::st_as_sf() |>            # convert from spatvector to sf
    st_intersection(plot_square_rastcut)  #crop to square plot

  ################################################
  # Crop observed ponds to the current square plot
  ponds_obs <- test_ponds |> 
    filter(dataset==dataset_sel) |> 
    st_intersection(st_geometry(plot_square)) |>
    mutate(pond_size = as.numeric(st_area(geometry))) |>   # add pond size
    mutate(rownum = row_number()) 

  ################################################
  # return metrics + obs/est datasets
  list(
    plot_id            = plot_square$Plot,
    ponds_obs_sf       = ponds_obs,
    ponds_est_sf       = ponds_est_sf,
    clouds_sf          = clouds_sf
    )
}
```

Fn to apply the above to different rasters/years
```{r}
get_test_results <- function(raster_list, dataset_map, waterval, cloudval) {

  # run for each dataset
  results <- lapply(names(raster_list), function(ds) {
  
    res_list <- lapply(
        which(test_plots_sf_square$Plot %in% test_plot_ids$Plot[test_plot_ids$dataset == ds] &
        test_plots_sf_square$dataset == ds),
      get_pond_metrics,
      raster      = raster_list[[ds]],
      dataset_sel = ds,
      waterval    = waterval,
      cloudval    = cloudval
    )
  
    list(
      dataset = ds,
      year    = dataset_map[[ds]],
      results = res_list
    )
  })

  # from results above, combine observed ponds sf
  ponds_obs_sf <- bind_rows(lapply(results, function(r) {
    do.call(bind_rows, lapply(r$results, function(x) {
      if (!is.null(x) && nrow(x$ponds_obs_sf) > 0) {
        cbind(
          plot_id = x$plot_id,
          year    = r$year,
          x$ponds_obs_sf
        )
      }
    }))
  })) |> mutate(type = "observed")

  # from results above, combine estimated ponds sf
  ponds_est_sf <- bind_rows(lapply(results, function(r) {
    do.call(bind_rows, lapply(r$results, function(x) {
      if (!is.null(x) && nrow(x$ponds_est_sf) > 0) {
        cbind(
          plot_id = x$plot_id,
          year    = r$year,
          x$ponds_est_sf
        )
      }
    }))
  })) |> mutate(type = "estimated")
  
  # from results above, combine clouds sf
  clouds_sf <- do.call(rbind, lapply(results, function(r) {
    do.call(rbind, lapply(r$results, function(x) {
      if (!is.null(x$clouds_sf) && nrow(x$clouds_sf) > 0) {
        x$clouds_sf$plot_id <- x$plot_id
        x$clouds_sf$year    <- r$year
        x$clouds_sf
      }
    }))
  }))
  
  # return clean object
  list(
    ponds_obs_sf = ponds_obs_sf,
    ponds_est_sf = ponds_est_sf,
    clouds_sf  = clouds_sf
  )
}
```

Apply functions
```{r}
model_pp1 <- get_test_results(
  raster_list = list("2022" = may2022_pp1, "2024" = may2024_pp1, "brood16" = aug16_pp1),
  dataset_map = c("2022" = "2022", "2024" = "2024", "brood16" = "brood16"),
  cloudval = 3,
  waterval = 1) 

dswe_agg <- get_test_results(
  raster_list = list("2022" = dswe_agg_May22, "2024" = dswe_agg_May24, "brood16" = dswe_agg_Aug16),
  dataset_map = c("2022" = "2022", "2024" = "2024", "brood16" = "brood16"),
  cloudval = 2,
  waterval = 1)

dswe_cons <- get_test_results(
  raster_list = list("2022" = dswe_cons_May22, "2024" = dswe_cons_May24, "brood16" = dswe_cons_Aug16),
  dataset_map = c("2022" = "2022", "2024" = "2024", "brood16" = "brood16"),
  cloudval = 2,
  waterval = 1)

dw <- get_test_results(
  raster_list = list("2022" = dw_May22, "2024" = dw_May24, "brood16" = dw_Aug16),
  dataset_map = c("2022" = "2022", "2024" = "2024", "brood16" = "brood16"),
  cloudval = 2,
  waterval = 1)

jrc <- get_test_results(
  raster_list = list("brood16" = jrc_Aug16),
  dataset_map = c("brood16" = "brood16"),
  cloudval = 0,
  waterval = 2)
```

Combine and export all polygons: observed and each model's estimations of water (plus clouds)
```{r}
# observed ponds
ponds_obs_sf <- model_pp1$ponds_obs_sf |> st_transform(4326)

# our model's estimated ponds and clouds
ponds_est_sf <- model_pp1$ponds_est_sf |> mutate(type="estimated_ours") |> st_transform(4326)
clouds_est <- model_pp1$clouds_sf  |> mutate(type="clouds_ours") |> st_transform(4326)

# dswe-conservative's estimated ponds and clouds
ponds_est_sf_cons <- dswe_cons$ponds_est_sf |> mutate(type="estimated_cons")
clouds_sf_cons <- dswe_cons$clouds_sf  |> mutate(type="clouds_cons")

# dswe-aggressive's estimated ponds and clouds
ponds_est_sf_agg <- dswe_agg$ponds_est_sf |> mutate(type="estimated_agg")
clouds_sf_agg <- dswe_agg$clouds_sf  |> mutate(type="clouds_agg")

# dynamic world's estimated ponds and clouds
ponds_est_sf_dw <- dw$ponds_est_sf |> mutate(type="estimated_dw")
clouds_sf_dw <- dw$clouds_sf |> mutate(type="clouds_dw")

# jrc's estimated ponds and clouds
ponds_est_sf_jrc<- jrc$ponds_est_sf |> mutate(type="estimated_jrc")
#clouds_sf_jrc <- jrc$clouds_sf |> mutate(type="clouds_jrc") # no clouds

all_sf <- bind_rows(
  ponds_obs_sf,
  ponds_est_sf,
  ponds_est_sf_cons,
  ponds_est_sf_agg,
  ponds_est_sf_dw,
  ponds_est_sf_jrc,
  clouds_est,
  clouds_sf_cons,
  clouds_sf_agg,
  clouds_sf_dw
) |> 
  select(-c(plot_id, year, Shp_Lng, Shap_Ar, utm_epsg, remapped, water))

st_write(all_sf, here("data/accuracy_assessment/all_est_obs_clouds/all_est_obs_clouds.shp"), append=F)
```