---
title: "Table S3: pixel accuracy where NDVI < 0"
author: "Maggie Church"
created: "2025-06-06"
updated: "2026-03-09"
output: html_document
---

```{r}
library(tidyverse)
library(here)
```

```{r}
source(here("code/analysis/make_figures/helper_functions/prep_pixel_results.R"))
source(here("code/analysis/make_figures/helper_functions/create_grouped_stats_table.R"))
source(here("code/analysis/make_figures/helper_functions/calculate_grouped_accuracy.R"))
```

Water is way less vegetated in Coteau, way more vegetated in Lake Agassiz (also just way less water sampled in Lake Agassiz and way more water in Coteau)
```{r}
pixel_results |>
  filter(!is.na(NDVI)) |>  # huh, surprising amount of missing NDVI... 
  group_by(ecoregion_name, true_label) |> 
  summarize(n_bare = sum(NDVI < 0), 
            n = n(), 
            p_bare = round(n_bare/n*100),
            mean_ndvi = round(mean(NDVI), 2),
            median_ndvi = round(median(NDVI), 2)) |> 
  arrange(desc(p_bare))
```

Get pixel metrics by ecoregion, where NDVI < 0
```{r}
metric_pixel_modelcomparison <- c("oa", "oe", "ce", "f1")
metric_pixel_modelcomparison_labels <- c("OA", "OE", "CE", "F1")

ts3 <- pixel_results |> 
  filter(NDVI < 0) |> 
  calculate_grouped_accuracy(group_cols = c("ecoregion_name"), 
                             model = "postproc10") |>
  arrange(ecoregion_name) |> 
  create_grouped_stats_table(group_col = ecoregion_name, 
                             model="postproc10",
                             metrics = metric_pixel_modelcomparison,
                             include_plot_accuracy = FALSE
                             ) 

ts3
```

```{r}
write_csv(ts3, here("figures/R/Table_S3.csv"))
```